% !TEX program = xelatex
%
% Epiphany --- Operation Catalog (companion specification)
% Companion to the Core Specification. Compile with XeLaTeX.
%
% This document is versioned independently of the Core Specification
% (independent semver; see the Versioning note in the front matter). Its preamble
% is intentionally a self-contained copy of the core specification's preamble so
% the two documents build independently; factoring a shared preamble file is a
% later cleanup, not a v0.1 deliverable.

\documentclass[11pt,letterpaper]{report}

% ---------------------------------------------------------------------------
% Packages
% ---------------------------------------------------------------------------
\usepackage{fontspec}
\usepackage{geometry}
\geometry{
  letterpaper,
  top=1.05in,
  bottom=1.05in,
  left=1.15in,
  right=1.15in,
  headheight=15pt
}

\usepackage[english]{babel}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tcolorbox}
\tcbuselibrary{breakable, skins}

% ---------------------------------------------------------------------------
% Color palette (shared with the core specification)
% ---------------------------------------------------------------------------
\definecolor{epiphanyteal}{HTML}{1A4044}
\definecolor{epiphanygold}{HTML}{8E6E2E}
\definecolor{epiphanyink}{HTML}{1F1B16}
\definecolor{epiphanyslate}{HTML}{6B6660}
\definecolor{epiphanycream}{HTML}{F8F4ED}
\definecolor{epiphanymist}{HTML}{ECE8E0}
\definecolor{epiphanycode}{HTML}{2A2520}
\definecolor{epiphanycrimson}{HTML}{7A2424}

\hypersetup{
  colorlinks=true,
  linkcolor=epiphanyteal,
  citecolor=epiphanyteal,
  urlcolor=epiphanygold,
  pdftitle={Epiphany --- Operation Catalog},
  pdfauthor={The Epiphany Project},
  pdfsubject={Operation Catalog companion for the Epiphany music notation platform},
  pdfkeywords={music notation, operations, CRDT, reduction, serialization},
  bookmarksnumbered=true,
  bookmarksopen=true
}

% ---------------------------------------------------------------------------
% Typography (shared with the core specification)
% ---------------------------------------------------------------------------
\setmainfont{TeX Gyre Pagella}[Numbers={OldStyle, Proportional}, Ligatures={TeX, Common}]
\setsansfont{TeX Gyre Heros}[Scale=0.94, Ligatures={TeX, Common}]
\setmonofont{TeX Gyre Cursor}[Scale=0.88, Ligatures={TeX}]
\newfontfamily\titlefont{TeX Gyre Pagella}[Numbers={OldStyle}, Ligatures={TeX, Common}]
\newcommand{\tablenums}[1]{{\addfontfeatures{Numbers={Lining,Tabular}}#1}}
\newcommand{\sectionsc}[1]{{\addfontfeatures{Letters=SmallCaps}#1}}

% ---------------------------------------------------------------------------
% Section styling (shared with the core specification)
% ---------------------------------------------------------------------------
\titleformat{\chapter}[display]
  {\normalfont\filright}
  {\raggedright\color{epiphanygold}\fontsize{14pt}{16pt}\selectfont
    \scshape Chapter\ \thechapter}
  {16pt}
  {\raggedright\color{epiphanyteal}\fontsize{32pt}{36pt}\selectfont\bfseries}
  [\vspace{4pt}{\color{epiphanygold}\rule{2in}{0.6pt}}]
\titlespacing*{\chapter}{0pt}{-20pt}{30pt}
\titleformat{\section}
  {\normalfont\Large\bfseries\color{epiphanyteal}}
  {\color{epiphanygold}\thesection}{1em}{}
\titleformat{\subsection}
  {\normalfont\large\bfseries\color{epiphanyteal}}
  {\color{epiphanygold}\thesubsection}{1em}{}
\titleformat{\subsubsection}
  {\normalfont\normalsize\bfseries\color{epiphanyink}}
  {\thesubsubsection}{1em}{}

% ---------------------------------------------------------------------------
% Headers and footers (shared with the core specification)
% ---------------------------------------------------------------------------
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\fancyhead[L]{\small\scshape\color{epiphanyslate}Epiphany --- Operation Catalog}
\fancyhead[R]{\small\itshape\color{epiphanyslate}\leftmark}
\fancyfoot[C]{\small\color{epiphanyslate}\thepage}
\renewcommand{\headrule}{
  \color{epiphanygold!50}\hrule width\headwidth height 0.4pt
  \vspace{1pt}
  \color{epiphanygold!30}\hrule width\headwidth height 0.2pt
}

% ---------------------------------------------------------------------------
% Code listing style (shared with the core specification)
% ---------------------------------------------------------------------------
\lstdefinelanguage{Rust}{
  keywords={fn,let,mut,pub,struct,enum,impl,trait,for,in,if,else,match,return,
           use,mod,crate,self,Self,as,where,move,async,await,const,static,
           ref,type,unsafe,extern,dyn,box,break,continue,loop,while},
  keywordstyle=\color{epiphanyteal}\bfseries,
  ndkeywords={i8,i16,i32,i64,i128,u8,u16,u32,u64,u128,f32,f64,bool,char,str,
              String,Vec,Option,Result,Box,Rc,Arc,HashMap,BTreeMap,
              NonZeroU16,NonZeroU32,NonZeroU64,Duration,Timestamp},
  ndkeywordstyle=\color{epiphanygold}\bfseries,
  sensitive=true,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{epiphanyslate}\itshape,
  stringstyle=\color{epiphanycrimson},
  morestring=[b]",
  morestring=[b]'
}
\lstset{
  basicstyle=\ttfamily\small\color{epiphanycode},
  backgroundcolor=\color{epiphanycream},
  frame=leftline,
  rulecolor=\color{epiphanygold!60},
  framesep=8pt,
  framerule=1.5pt,
  xleftmargin=10pt,
  xrightmargin=4pt,
  breaklines=true,
  showstringspaces=false,
  numberstyle=\tiny\color{epiphanyslate},
  numbersep=10pt,
  captionpos=b,
  aboveskip=10pt,
  belowskip=10pt,
  language=Rust
}

% ---------------------------------------------------------------------------
% Custom environments (shared with the core specification)
% ---------------------------------------------------------------------------
\newtcolorbox{openquestion}[1][]{
  enhanced, breakable,
  colback=epiphanymist, colframe=epiphanycrimson,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Open Question},
  coltitle=white, colbacktitle=epiphanycrimson,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}
\newtcolorbox{rationale}[1][]{
  enhanced, breakable,
  colback=epiphanymist, colframe=epiphanyteal,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Rationale},
  coltitle=white, colbacktitle=epiphanyteal,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}
\newtcolorbox{requirement}[1][]{
  enhanced, breakable,
  colback=white, colframe=epiphanygold,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Requirement},
  coltitle=white, colbacktitle=epiphanygold,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}
\newtcolorbox{nongoal}[1][]{
  enhanced, breakable,
  colback=epiphanymist, colframe=epiphanyslate,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Non-Goal},
  coltitle=white, colbacktitle=epiphanyslate,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}

\newcommand{\MUST}{\textbf{MUST}}
\newcommand{\MUSTNOT}{\textbf{MUST}\nobreak\ \textbf{NOT}}
\newcommand{\SHOULD}{\textbf{SHOULD}}
\newcommand{\SHOULDNOT}{\textbf{SHOULD}\nobreak\ \textbf{NOT}}
\newcommand{\MAY}{\textbf{MAY}}

\setlist[itemize]{topsep=2pt, itemsep=3pt, parsep=0pt}
\setlist[enumerate]{topsep=2pt, itemsep=3pt, parsep=0pt}
\setlist[description]{topsep=2pt, itemsep=5pt, parsep=0pt}
\AtBeginDocument{\color{epiphanyink}}

% ---------------------------------------------------------------------------
% Document
% ---------------------------------------------------------------------------
\begin{document}

\begin{titlepage}
  \thispagestyle{empty}
  \centering
  \vspace*{2.2in}
  {\color{epiphanygold}\rule{3in}{0.8pt}}\\[18pt]
  {\titlefont\fontsize{34pt}{38pt}\selectfont\color{epiphanyteal}\bfseries Epiphany}\\[10pt]
  {\Large\scshape\color{epiphanyslate}Operation Catalog}\\[6pt]
  {\large\itshape\color{epiphanyslate}A companion to the Core Specification}\\[14pt]
  {\color{epiphanygold}\rule{3in}{0.8pt}}\\[24pt]
  {\normalsize\color{epiphanyink}Version 0.8.0 --- Transpose algebra (TransposeInterval; the frozen Transpose)}\\[4pt]
  {\small\color{epiphanyslate}Normative for the operation kinds it defines}
  \vfill
\end{titlepage}

\tableofcontents

% ===========================================================================
\chapter{About This Companion}
\label{ch:about}

The \emph{Operation Catalog} is a companion to the Epiphany Core Specification.
It fulfils the open question the core specification raises in its
\emph{Operation Catalog Conformance} section (Chapter~6,
\sectionsc{Semantic Operations and Concurrent Reduction}, the
\texttt{sec:semops:catalog} open question), which states that the catalog
``is normative once published; this specification is non-final until the catalog
is delivered.'' This release delivers the catalog \emph{framework}, the
\textbf{K0 representative primitive set}, and the \textbf{M2 broad-K0 groups}
(the event/pitch leaf-field, cross-cutting CRUD, structural-container, and
score-settings operations the Phase~2 slice exercises) --- the operation kinds
the Phase~2 visible slice and binary format actually exercise, each fully
specified in Chapter~\ref{ch:k0}. The remaining items of the full
$60$--$80$-primitive catalog are drafted as framework slots
(Chapter~\ref{ch:k1}) and completed in Phase~3.

\section{Relationship to the Core Specification}

This companion does not restate the operation framework; it \emph{references} it.
The framework --- operation identity and stamps, the hybrid-logical-clock
monotonicity rule, the dotted-version-vector causal context, the
order-independent operation-slot model and equivocation, the canonical reduction
order, the four-phase lifecycle, conflict records and the conflict registry,
re-anchoring, transactions, forward undo, and the LWW discipline --- is the core
specification's Chapter~6. This companion defines, for each operation kind, the
\emph{payload schema} and how that kind \emph{instantiates} the framework's
reduction, conflict, undo, and re-anchoring rules.

It also consumes, rather than re-deriving, the \textbf{ratified byte-convention
baseline} (core specification Chapter~8, \sectionsc{Binary Format Companion},
requirement \texttt{req:format:codec-conventions}, and the
\sectionsc{Canonical Byte-Layout Reference} appendix): little-endian integers, a
single discriminant byte per tagged union, \texttt{u32} length prefixes on every
variable-width leaf, and raw UTF-8 free text. Operation-payload encodings are
expressed in terms of that baseline; the literal wire layout of each payload is
the Binary Format companion's (Agent~J's) to pin, in coordination with this
catalog.

\begin{rationale}
The catalog is versioned \emph{independently} of the core specification
(independent semver). Operation kinds are added over time; the catalog should
evolve --- adding primitives, refining conflict cases --- without forcing a core
specification revision. The core specification changes only when the
\emph{framework} changes.
\end{rationale}

\section{Conformance Profiles}

A \textbf{Phase-2 profile} implementation \MUST{} implement every primitive in
Chapter~\ref{ch:k0} --- the representative set and the M2 broad-K0 groups --- with
the schema, reduction rule, conflict cases, undo semantics, and re-anchoring
behaviour defined there. The former create-score/create-canvas framework slots
of Chapter~\ref{ch:k1} are \emph{retired} (Pass~12): they are permanently
outside the operation set, so no operation kind exists to reject.

\medskip

\noindent\textbf{Version 0.6.0 (Pass-12 G-ratification).} No payload byte
changes. Ratified: ResolveConflict beyond the concurrent case (P12-K4),
ResolveEquivocation edge semantics (P12-K6) and the profile-policy deferral
(P12-K5), the RespellPitch migration fallback as long-term (P12-K1), undo
strand-block conflict-kind reuse (P12-K10) and the idempotence asymmetry
(P12-K11), the Transpose prototype pin (P12-K2 deferral), the cross-region
slur AND advisory (P12-K12), and the create-score/canvas retirement (P12-K8).
Appended vocabulary (minor, append-only):
\texttt{SystemDerivedContentImmutable} (12, P12-K3) and
\texttt{RecreateContentMismatch} (13, P12-K9) in
\texttt{PreconditionFailureReason}.

\medskip

\noindent\textbf{Version 0.8.0 (Push~4a, transpose algebra).} Adds one
operation kind, \texttt{TransposeInterval}
(Section~\ref{sec:k0:transpose-interval}), carrying the existing
\texttt{TranspositionInterval} of core Chapter~2. It transposes scale
position, refuses
atomically rather than saturating, takes a genuine \emph{set} of targets, and
propagates the resulting spelling. \texttt{Transpose}
(Section~\ref{sec:k0:transpose}) is \textbf{frozen}: its saturation, its
nominal-blindness, and its duplicate-sensitivity are pinned as normative
replay semantics and no longer authored. This supersedes the 0.6.0 pin of the
Transpose prototype (P12-K2), which anticipated a payload schema-major landing
with the tuning catalog; neither proved necessary. Appended vocabulary (minor,
append-only): \texttt{AcousticRealizationPinned} (14) and
\texttt{TranspositionOutOfRange} (15) in \texttt{PreconditionFailureReason};
\texttt{PitchSpaceMismatch} (6) is un-reserved and now produced, its previous
dependency on the tuning catalog having been an error --- detecting a
non-\texttt{Cmn} position reads a discriminant, not a pitch-space registry.

% ===========================================================================
\chapter{The Catalog Framework}
\label{ch:framework}

\section{Value-Typed Payloads}
\label{sec:framework:value-typed}

Every operation payload in this catalog is \textbf{value-typed}: it carries the
real graph values its effect introduces, not identifiers that point at values in
some ambient graph. An \texttt{InsertEvent} carries the whole \texttt{Event}; a
\texttt{RespellPitch} carries the whole \texttt{PitchSpelling}; a
\texttt{CreateCrossCutting} carries the whole tie, slur, beam, or spanner.

\begin{rationale}
The v0 prototype carried \emph{identifier-only projections} --- an
\texttt{InsertEvent} held an \texttt{EventId} plus reduction-relevant scalars; a
\texttt{RespellPitch} held a content-hash \emph{fingerprint} of the new spelling.
That was sufficient to make an envelope hashable and to drive reduction, but it
is \textbf{not durable}: an operation replayed in a fresh context --- a backup
restore, a cross-tool round-trip --- needs the full value, which an
identifier-only payload cannot supply. Value-typed payloads are what make an
Epiphany document portable.
\end{rationale}

\begin{requirement}
\label{req:catalog:value-encoding}
A value-typed payload field \MUST{} be encoded by emitting the field's value
under the core specification's canonical value encoding
(\texttt{req:format:codec-conventions}), framed by a \texttt{u32} little-endian
length prefix. Decoding \MUST{} be the exact inverse and \MUST{} reject trailing
bytes within the framed region. The encoding introduces no byte layout beyond the
ratified baseline: a value's bytes here are byte-for-byte the bytes the whole
document codec emits for that value.
\end{requirement}

\section{Per-Primitive Schema Template}
\label{sec:framework:template}

Each catalog primitive is specified under a fixed six-part template. A new
primitive is a schema-fill against this template, not a fresh design:

\begin{description}
  \item[Payload schema] The value-typed fields the operation carries, with their
    core-specification types.
  \item[Canonical encoding] The field order and framing, consuming
    Requirement~\ref{req:catalog:value-encoding}. (The literal byte layout is the
    Binary Format companion's; this catalog fixes the \emph{field set and order}.)
  \item[Reduction rule] How the operation mutates canonical state when it is
    reached in canonical reduction order --- the preconditions it checks, the
    objects it mints or tombstones, and the bookkeeping it records.
  \item[Conflict cases] The conflict records the operation can produce, by
    \texttt{ConflictKind}, and which participant materialises.
  \item[Undo semantics] The compensating effect of undoing a transaction that
    contains the operation, under each \texttt{UndoPolicy}
    (\texttt{StrictInverse} / \texttt{BestEffort} / \texttt{Cascade}).
  \item[Re-anchoring] The behaviour when an object the operation references is
    tombstoned before or concurrently with it.
\end{description}

\section{Reduction-Discipline Coverage}

The K0 representative set is chosen so that, between them, the primitives
exercise \emph{every} reduction discipline the framework defines:
position-keyed insert with system-voice promotion; delete-wins with tombstones,
tuplet compensation, and cross-cutting re-anchoring; field-overwrite with
last-writer-wins and structural-field-collision conflicts; set-union creation;
structural time-model migration; an LWW advisory; atomic transactions with
descriptor precedence; and the two meta-operations (conflict resolution and
forward undo). A primitive added later that reuses one of these disciplines
inherits its reduction, conflict, undo, and re-anchoring treatment.

% ===========================================================================
\chapter{K0 --- Representative Primitives}
\label{ch:k0}

This chapter is normative under the Phase-2 profile. Each primitive's reduction,
conflict, undo, and re-anchoring behaviour is the behaviour the core
specification's Chapter~6 defines for its discipline; the description here states
how the primitive instantiates it. The reference implementation is
\texttt{epiphany-ops} (the \texttt{payload}, \texttt{reduce}, and \texttt{migrate}
modules).

\section{InsertEvent}
\label{sec:k0:insert-event}

\textbf{Payload schema.} \texttt{InsertEventOp \{ staff\_instance:
StaffInstanceId, event: Event \}}. The voice, region-local position, duration,
and pitch identities are read from the \texttt{Event} value; the
\texttt{staff\_instance} is retained alongside it so the system-voice promotion
derivation is total without a containment walk.

\textbf{Canonical encoding.} \texttt{staff\_instance}, then the length-framed
canonical bytes of \texttt{event}.

\textbf{Reduction rule.} A position-keyed insert. Preconditions: the event's
duration is positive; the event id is neither live nor tombstoned; in graph-aware
reduction the target voice exists in a metric region and the event's pitch ids
are fresh. The event and its pitches are minted live; the voice is created on
first use. Concurrent inserts whose half-open duration intervals overlap in the
same voice are resolved by an order-independent promotion pre-pass: the
lower-\texttt{OperationId} insert is retained in the original voice, and each
overlapping loser is promoted to the deterministic system voice
\texttt{derive\_promoted\_voice\_id(staff\_instance, voice, winner, loser)} and
tagged \texttt{VoicePromoted}.

\textbf{Conflict cases.} None at reduction time; promotion is a deterministic
repair, not a conflict.

\textbf{Undo semantics.} Undoing the enclosing transaction tombstones the minted
event, its pitches, and any promoted voice. \texttt{StrictInverse} conflicts if
any was already tombstoned or modified; \texttt{BestEffort} tombstones the
survivors; \texttt{Cascade} is \texttt{StrictInverse} over the same minted set
(dependent-closure undo is a Phase-3 refinement).

\textbf{Re-anchoring.} Not applicable (the operation mints, it does not
reference a pre-existing object that could be tombstoned).

\section{DeleteEvent}
\label{sec:k0:delete-event}

\textbf{Payload schema.} \texttt{DeleteEventOp \{ event: EventId,
tuplet\_compensation: TupletCompensation \}}, where \texttt{TupletCompensation}
is \texttt{NotInTuplet}, \texttt{ReplaceWithRest \{ rest: Rest \}} (value-typed
replacement rest), \texttt{RewriteTuplets \{ tuplets \}}, or
\texttt{CascadeDeleteTuplets \{ tuplets \}}.

\textbf{Canonical encoding.} \texttt{event}, then the tuplet-compensation
discriminant and its payload (a length-framed \texttt{Rest} value for
\texttt{ReplaceWithRest}).

\textbf{Reduction rule.} Delete-wins: the event and its contained pitches are
tombstoned (retaining their identifiers). Tuplet compensation, when present, adds
the replacement rest live or tombstones the cascaded tuplet group. Concurrent
deletes of the same event are idempotent.

\textbf{Conflict cases.} None for the delete itself. Graph-aware reduction
refuses an ill-formed tuplet compensation as a precondition failure (no-op).

\textbf{Undo semantics.} An insert-shaped compensation re-introduces the
tombstoned content; for the prototype's minted-object model this is the inverse
of the tombstone set, with the policy treatment described under InsertEvent.

\textbf{Re-anchoring.} Tombstoning the event runs the framework's re-anchoring
rule table over every cross-cutting structure that referenced it: a tie
cascade-deletes; a comment or analytical annotation orphans (user content is
never silently deleted); a beam truncates while $\geq 2$ members survive and
otherwise cascade-deletes; a slur or spanner re-anchors to the nearest surviving
endpoint while $\geq 1$ survives and otherwise cascade-deletes.

\section{RespellPitch}
\label{sec:k0:respell-pitch}

\textbf{Payload schema.} \texttt{RespellPitchOp \{ pitch: PitchId, spelling:
PitchSpelling \}} --- the full spelling value (v1), not a fingerprint.

\textbf{Canonical encoding.} \texttt{pitch}, then the length-framed canonical
bytes of \texttt{spelling}.

\textbf{Reduction rule.} A last-writer-wins field overwrite, keyed by pitch.
Precondition: the pitch is live. The resolved spelling is the one carried by the
operation latest in canonical order. Two respellings of one pitch that are
causally ordered overwrite intentionally; two \emph{concurrent} respellings with
\emph{equal} spelling reduce idempotently.

\textbf{Conflict cases.} Two concurrent respellings of one pitch with
\emph{differing} spellings produce a \texttt{StructuralFieldCollision} conflict
recording the winner (later in canonical order), the loser, and the field
\texttt{spelling}. The winner materialises and carries the \texttt{Conflicted}
effect tag.

\textbf{Undo semantics.} Value-restoring (Section~\ref{sec:k0:undo}): undo
restores the pre-operation spelling (or removes the spelling if the operation
introduced the first one), under the active policy.

\textbf{Re-anchoring.} If the target pitch is tombstoned, the respelling is a
no-op (\texttt{TargetTombstoned}).

\emph{Migration (ratified Pass~12, closing P12-K1).} A v0
\texttt{RespellPitch} carried only a content-hash \emph{fingerprint} of the
spelling. The fingerprint cannot be inverted to a \texttt{PitchSpelling}
without a side table, so the v0$\rightarrow$v1 migration
(Chapter~\ref{ch:migration}) recovers the spelling from the score graph
context --- an explicit per-pitch spelling attachment whose canonical bytes
hash to the fingerprint --- and, when the context lacks it, declares the
envelope unmigratable (the bundle opens read-only). This read-only fallback
is the \emph{long-term} disposition: no richer v0 corpus is or will be
required (no production v0 corpus exists). It remains the one representative
payload that is not self-contained under migration.

\section{ModifyEvent}
\label{sec:k0:modify-event}

\textbf{Payload schema.} \texttt{ModifyEventOp \{ event: Event \}} --- the full
replacement \texttt{Event} value (v1). The identity (and therefore the LWW key)
is read from the value.

\textbf{Canonical encoding.} The length-framed canonical bytes of \texttt{event}.

\textbf{Reduction rule.} A last-writer-wins field overwrite keyed by event id.
Precondition: the event is live. The resolved value is the one carried by the
operation latest in canonical order; two causally ordered modifications overwrite
intentionally, and two \emph{concurrent} modifications with \emph{equal} value
reduce idempotently. Graph-aware reduction overwrites the event in place, and a
modification that \emph{moves} a metric event (a different region-local
\texttt{Musical} position or duration) is \emph{materialised}: the owning voice
is re-sorted by ascending position (id-tiebroken --- the same order an insert
maintains), preserving \texttt{VoiceEventsSortedNonOverlap} (Chapter~5
invariant~3). To keep that invariant, a placement change carries a
\emph{placement precondition}, read from the reducer's canonical voice-occupancy
index (graph-independent, so graph-free and graph-aware reduction agree): a move
with a non-positive span, or one that would overlap another live event in the
voice, is refused as a clean precondition no-op (\texttt{EventDurationInvalid})
rather than skipped silently. A materialised move updates the occupancy index,
so a later insert sees the freed or changed span. A placement change of a
\emph{non-metric} event is recorded in the bookkeeping but not applied to the
graph (re-sorting a non-metric voice is a deferred refinement); a malformed
(empty pitched) replacement is likewise recorded but not materialised;
same-placement field edits apply in place, preserving voice membership. Partial
trimming of a tuplet member remains a later refinement.

\emph{System-derived content immutability (ratified Pass~12, closing
P12-K3).} A modification that would rewrite the \emph{intrinsic content} of a
pitch whose identifier lives in the \texttt{SYSTEM\_DERIVED} namespace is
refused as a clean precondition no-op with the appended reason
\texttt{SystemDerivedContentImmutable} (discriminant 12): the identifier is
content-derived, and an in-place rewrite would silently invalidate its
derivation (core specification, Chapter~5 system-derived identity). The same
precondition applies to \texttt{ModifyIdentifiedPitch}
(Section~\ref{sec:k0:identified-pitch}). The sanctioned path is minting a
replacement pitch.

\textbf{Conflict cases.} Two concurrent modifications of one event with
\emph{differing} values produce a \texttt{StructuralFieldCollision} on the field
\texttt{event}, recording the winner (later in canonical order) and the loser.

\textbf{Undo semantics.} Value-restoring (Section~\ref{sec:k0:undo}): undoing
the enclosing transaction restores the event's chain-predecessor value,
conflicting (\texttt{StrictInverse}) or skipping (\texttt{BestEffort}) when a
later modification has superseded it.

\textbf{Re-anchoring.} If the target event is tombstoned, the modification is a
no-op (\texttt{TargetTombstoned}/\texttt{TargetMissing}).

\section{Identified-Pitch Operations}
\label{sec:k0:identified-pitch}

\textbf{Payload schema.} \texttt{InsertIdentifiedPitchOp \{ event: EventId,
pitch: IdentifiedPitch \}} mints a pitch into a live event;
\texttt{DeleteIdentifiedPitchOp \{ pitch: PitchId \}} tombstones one;
\texttt{ModifyIdentifiedPitchOp \{ pitch: PitchId, value: Pitch \}} overwrites a
pitch's acoustic / scale-position value (distinct from \texttt{RespellPitch},
which overwrites only the \emph{spelling}).

\textbf{Canonical encoding.} Insert: \texttt{event}, then the length-framed
\texttt{pitch} value. Delete: \texttt{pitch}. Modify: \texttt{pitch}, then the
length-framed \texttt{value}.

\textbf{Reduction rule.} The pitch-level analogues of the event-level mint,
delete, and field overwrite, inheriting their disciplines
(Sections~\ref{sec:k0:insert-event}, \ref{sec:k0:delete-event}, and this
chapter's field-overwrite treatment). \textbf{A note and a rest are the same slot
under pitch add/remove} (normative): deleting the \emph{only} pitch of a
single-pitch note degrades the event to a \texttt{Rest} of the same
id/voice/position/duration rather than leaving an empty pitched event (Chapter~5
forbids the empty chord, \texttt{ArenaError::EmptyPitchedEvent}); inserting a
pitch into a rest is the dual, promoting it to a one-pitch note. This preserves
the delete-wins / mint disciplines and keeps the graph consistent with the
bookkeeping, which tombstones or mints the pitch object either way.

\textbf{Conflict cases.} Insert and delete: none (mint is set-union; delete-wins
is idempotent). Modify: two concurrent differing writes of one pitch produce a
\texttt{StructuralFieldCollision} on the field \texttt{pitch}.

\textbf{Undo semantics.} Insert: undoing the enclosing transaction tombstones
the minted pitch, re-resting the event if it was the only one
(Section~\ref{sec:k0:undo}). Modify: value-restoring
(Section~\ref{sec:k0:undo}) --- the pitch's chain-predecessor value is
restored unless superseded. Delete: still not inverted (re-introducing a
tombstoned pitch is the deferred resurrection case, P11-C8).

\textbf{Re-anchoring.} An operation whose target event or pitch is tombstoned is
a no-op; tombstoning a pitch runs the cross-cutting re-anchoring table over any
structure that referenced it (see DeleteEvent).

\section{Transpose (frozen; replay only)}
\label{sec:k0:transpose}

\begin{requirement}
  \label{req:opcat:transpose-frozen}
  \texttt{Transpose} is \textbf{frozen}. Its reduction rule is fixed at the
  semantics described in this section, including the saturation and the
  duplicate-sensitivity, for as long as any stored operation carries it. A
  conforming implementation \MUST{} reduce it exactly as specified here and
  \MUSTNOT{} emit it from new authoring; new transpositions are authored as
  \texttt{TransposeInterval} (Section~\ref{sec:k0:transpose-interval}).
\end{requirement}

\textbf{Payload schema.} \texttt{TransposeOp \{ targets: Vec<PitchId>,
chromatic\_steps: i32 \}}. Pitch identifiers are preserved. The operation
rewrites \emph{scale position} (the \texttt{Cmn} alteration), not acoustic
content.

\textbf{Canonical encoding.} \texttt{targets} in ascending canonical-byte order,
then \texttt{chromatic\_steps} as a little-endian \texttt{i32}. The sequence is
\emph{sorted but not deduplicated}: it is a multiset on the wire.

\textbf{Reduction rule.} An order-dependent content overwrite. For each live,
non-system target \emph{occurrence} in \texttt{targets}, the target's \texttt{Cmn}
\texttt{alteration} is shifted by \texttt{chromatic\_steps} and clamped to the
\texttt{i8} bound. Consequences, all normative for replay:

\begin{itemize}
  \item A target appearing $k$ times is shifted $k$ times.
  \item The nominal and octave never change: transposing C4 by $+12$ yields C4
    with \texttt{alteration} $= 12$, not C5.
  \item A shift past the \texttt{i8} bound saturates silently, and the operation
    still reports \texttt{Applied}. Transposition under this rule is therefore
    not invertible.
  \item A target whose position is not \texttt{Cmn} is left unchanged, and the
    operation still reports \texttt{Applied}.
\end{itemize}

\begin{rationale}
  Every bullet above is a defect (P12-K2), and none of them may be repaired in
  place. An operation is history: a replica replaying a stored
  \texttt{Transpose} must reconstruct the state its author saw, and a
  ``corrected'' reduction rule would silently rewrite every score that ever
  used one. The defects are therefore \emph{pinned}, not fixed, and the repair
  ships as a new operation kind whose discriminant no historical operation
  carries. Ratified Push~4a; supersedes the Pass~12 pin, which anticipated a
  payload schema-major and a dependency on the tuning catalog. Neither proved
  necessary: transposition acts on scale position, and appending an operation
  kind is a schema \emph{minor} (Binary Format companion, requirement
  \texttt{req:binfmt:kind-discriminants}).
\end{rationale}

\textbf{Conflict cases.} None --- composition is deterministic in canonical order
(a deterministic repair, not a conflict).

\textbf{Undo semantics.} Transpose mints nothing, so the prototype's minted-object
undo (Section~\ref{sec:k0:undo}) does not negate it, and it records no write into
the pitch's value chain, so \emph{value-restoring undo does not restore it
either}: undoing a transaction containing a \texttt{Transpose} leaves the shifted
pitch shifted. An inverse-interval undo does not exist (under saturation the
inverse is not a function) and remains a deferred refinement (P11-C8).

\textbf{This too is frozen.} Making \texttt{Transpose} record into the value
chain would not change its own reduction rule, but it \emph{would} change what a
stored history $\{\texttt{Transpose}, \texttt{UndoTransaction}\}$ replays to ---
from "the pitch stays shifted" to "the pitch returns" --- which is a change in
what an existing document means. \texttt{TransposeInterval} records its write and
\emph{is} undoable (Section~\ref{sec:k0:transpose-interval}); the frozen
operation is permanently not. That asymmetry is another reason never to author
it.

\textbf{Re-anchoring.} Tombstoned targets are skipped (the transpose applies only
to live pitches). \texttt{SYSTEM\_DERIVED}-namespace targets are likewise
\emph{skipped} (ratified Pass~12, P12-K3): their intrinsic content is immutable,
and an in-place alteration shift would desynchronize the content from the id's
derivation inputs. A transpose whose live targets are \emph{all} system-derived
reduces as a precondition no-op (\texttt{SystemDerivedContentImmutable}).

\section{TransposeInterval}
\label{sec:k0:transpose-interval}

\textbf{Payload schema.} \texttt{TransposeIntervalOp \{ targets:
CanonicalSet<PitchId>, interval: TranspositionInterval \}}, reusing the
diatonic/chromatic pair the core specification defines in Chapter~2,
\sectionsc{Transposition and the Interval Type} --- the same value an
\texttt{Instrument} carries as its written-versus-sounding interval. Pitch
identifiers are preserved; the operation rewrites scale position and leaves
acoustic realization untouched.

\begin{requirement}
  \label{req:opcat:transpose-interval-targets}
  \texttt{targets} is a \textbf{set}. It \MUST{} be encoded as a sequence in
  ascending canonical-byte order with \emph{strictly increasing} elements; a
  decoder \MUST{} reject a sequence containing a duplicate rather than
  normalize it. A transposition is a function of \emph{which} pitches it names,
  never of how many times it names them.
\end{requirement}

\textbf{Canonical encoding.} The strictly-increasing \texttt{targets} sequence,
then \texttt{interval.diatonic\_steps} and \texttt{interval.chromatic\_steps},
each a little-endian \texttt{i32}.

\textbf{Reduction rule.} An order-dependent content overwrite. Tombstoned and
\texttt{SYSTEM\_DERIVED} targets are skipped, exactly as for \texttt{Transpose};
if no mutable target remains the operation reduces as the corresponding
precondition no-op. Otherwise, every remaining target \MUST{} be transposable
under core specification requirement \texttt{req:pitch:transposition}, and each
is transposed by \texttt{interval}.

\begin{requirement}
  \label{req:opcat:transpose-interval-atomic}
  If any mutable target is \emph{not} transposable, the operation \MUST{}
  reduce to a no-op that changes no pitch, reporting the precondition failure
  of the first such target in canonical order. It \MUSTNOT{} transpose the
  remaining targets. The three cases and their
  \texttt{PreconditionFailureReason} discriminants:
  \begin{itemize}
    \item a non-\texttt{Cmn} \texttt{scale\_position.position} $\Rightarrow$
      \texttt{PitchSpaceMismatch} (6);
    \item an \texttt{AcousticRealization::AbsoluteHz} realization
      $\Rightarrow$ \texttt{AcousticRealizationPinned} (14);
    \item an \texttt{alteration}$'$ or \texttt{octave}$'$ that does not fit
      its field $\Rightarrow$ \texttt{TranspositionOutOfRange} (15).
  \end{itemize}
\end{requirement}

\begin{requirement}
  \label{req:opcat:transpose-interval-spelling}
  A \texttt{TransposeInterval} \MUST{} record the spelling its interval
  determined as a \texttt{SpellingAttachment} with source
  \texttt{SpellingSource::Propagated}, per the core specification's Chapter~2,
  \sectionsc{Spelling Sources}. The attachment's \texttt{from} is the
  transposed pitch's own identifier, which the operation preserves.

  That attachment alone is \emph{not sufficient}, because the default
  precedence ranks \texttt{UserChosen} and \texttt{Imported} above
  \texttt{Propagated} (Chapter~2, \sectionsc{Configurable Precedence}). A
  \texttt{TransposeInterval} \MUST{} therefore also \textbf{move} every
  engraved-layer, pitch-scoped, explicit spelling attachment on each target
  that is not itself \texttt{Propagated}, preserving each attachment's
  \texttt{source}, \texttt{priority}, and \texttt{layer}. A spelling is moved
  by transposing its \emph{nominal} diatonically and taking whatever accidental
  the transposed pitch then requires at that staff position. If any such
  attachment cannot be written at the transposed position, the operation
  \MUST{} refuse, atomically, with \texttt{TranspositionOutOfRange}.
\end{requirement}

\begin{rationale}
  Two different failures hide here, and the propagated attachment addresses only
  the first.

  For a pitch with \emph{no} authored spelling, the pre-pass would re-infer, and
  may pick the enharmonic the interval did not choose. The propagated attachment
  pins the interval's determination.

  For a pitch \emph{with} an authored spelling, the propagated attachment is
  outranked and changes nothing: a C4 the author spelled ``C'', sharpened to
  C-sharp~4, would still resolve to \texttt{Authored(UserChosen)} and engrave a
  C natural --- the accidental simply disappears, and the notehead names a pitch
  that is not there. So the authored spelling is moved instead of being left or
  discarded. It is moved by its \textbf{nominal}, because the nominal is what
  carries the author's enharmonic decision: an author who wrote B-sharp~3 rather
  than C4 chose the letter B, and a perfect fifth up must give F-double-sharp~4,
  not G. Discarding the attachment would silently lose that decision; re-inferring
  it from the transposed pitch would give G.

  A transposed \texttt{UserChosen} spelling is still the user's choice, so its
  source is preserved. \texttt{Imported} likewise: an imported attachment asserts
  ``this pitch is written thus'', and once the pitch moves that assertion must
  move with it or become false. Import fidelity is a property of the source file,
  which a transposition does not touch.
\end{rationale}

\textbf{Conflict cases.} None --- composition is deterministic in canonical order.

\textbf{Undo semantics.} Nothing is minted. \texttt{TransposeInterval}
\MUST{} record each transposed pitch's new value into that pitch's value chain,
and each spelling attachment it rewrote into the corresponding spelling chain, so
that value-restoring undo recovers the pre-transpose pitch \emph{and} its
pre-transpose spelling. Restoring the pitch alone would leave a notehead spelled
for a pitch that no longer exists. This is where \texttt{TransposeInterval}
departs from the frozen \texttt{Transpose}, which records nothing and is
therefore not undoable.

An inverse interval usually exists ($(-d, -c)$), because the reduction never
saturates --- but not always: $\texttt{i32::MIN}$ has no negation. Undo does not
rely on it.

\textbf{Re-anchoring.} As \texttt{Transpose}.

\section{CreateCrossCutting}
\label{sec:k0:create-cross-cutting}

\textbf{Payload schema.} \texttt{CreateCrossCuttingOp \{ structure:
CrossCuttingValue \}}, where \texttt{CrossCuttingValue} is the typed value of a
\texttt{Tie}, \texttt{Slur}, \texttt{Beam}, or \texttt{Spanner}. Its identity and
referenced endpoints are read from the value.

\textbf{Canonical encoding.} A discriminant for the structure kind, then the
length-framed canonical bytes of the structure value.

\textbf{Reduction rule.} Set-union creation: the structure is minted live if its
id is not already live and every referenced endpoint is live; a second create of
a live id is idempotent. Graph-aware reduction materialises the structure with
its full fields.

\textbf{Conflict cases.} None (all-or-nothing creation; union is deterministic).

\textbf{Undo semantics.} Undo tombstones the minted structure, under the active
policy.

\textbf{Re-anchoring.} The structure participates in the re-anchoring rule table
when one of its endpoints is later tombstoned (see DeleteEvent).

\textbf{Authoring advisory (ratified Pass~12, closing P12-K12).} The
slur-spanning advisory reads
\texttt{Region.permits\_spanning\_slurs} \emph{conjunctively}: a slur (or
other spanner) whose endpoints lie in different regions passes the advisory
only when \emph{both} endpoint regions permit spanning. The check is
authoring-time only and never alters reduction.

\textbf{Migration coverage.} The v0$\rightarrow$v1 migration
(Chapter~\ref{ch:migration}) reconstructs the event-anchored \texttt{Tie},
\texttt{Slur}, and \texttt{Beam} from the v0 reference (id plus event endpoints).
A \texttt{Spanner} is anchored by \texttt{TimeAnchor}s rather than a fixed pair
of event endpoints, so its full value is not reconstructable from the v0
event-reference projection; a \texttt{Spanner}-create is therefore reported
unmigratable (read-only), alongside the respell case of P12-K1, and remains so.
A faithful spanner migration awaits a richer v0 projection that carries the
anchors --- a Phase-3 / Pass-12 extension, not yet implemented.

\section{DeleteCrossCutting}
\label{sec:k0:delete-cross-cutting}

\textbf{Payload schema.} \texttt{DeleteCrossCuttingOp \{ structure: TypedObjectId
\}} --- the structure named by the same key the set-union creation and the
re-anchoring table use; it \MUST{} be a cross-cutting kind
(\texttt{Tie}/\texttt{Slur}/\texttt{Beam}/\texttt{Spanner}).

\textbf{Canonical encoding.} The canonical bytes of \texttt{structure}.

\textbf{Reduction rule.} Delete-wins: the structure is tombstoned (its identifier
retained). A second delete of the same structure is idempotent; deleting a
missing or non-cross-cutting id is a no-op precondition failure. Graph-aware
reduction removes the structure from the score.

\textbf{Conflict cases.} None (delete-wins is idempotent).

\textbf{Undo semantics.} A delete mints nothing, so the prototype's minted-object
undo (Section~\ref{sec:k0:undo}) does not re-introduce the tombstoned structure
(P11-C8).

\textbf{Re-anchoring.} The deletion is direct (the structure is the target, not a
referenced endpoint); it does not itself trigger the endpoint re-anchoring table.

\section{ModifyCrossCutting}
\label{sec:k0:modify-cross-cutting}

\textbf{Payload schema.} \texttt{ModifyCrossCuttingOp \{ structure:
CrossCuttingValue \}} --- the full replacement value (v1). The replacement keeps
the structure's identity but may change its endpoints and per-kind fields; the
LWW key is the structure's \texttt{CrossCuttingValue::id}.

\textbf{Canonical encoding.} The discriminant and length-framed bytes of the
\texttt{CrossCuttingValue} (as for CreateCrossCutting).

\textbf{Reduction rule.} A last-writer-wins field overwrite keyed by structure
id. Precondition: the structure is live. The reduction re-derives the structure's
endpoints from the new value, so a later re-anchoring sees them. A malformed
replacement is a precondition no-op --- in particular a beam whose membership
falls below the two-member minimum is refused rather than materialised.
Graph-aware reduction overwrites the structure in place.

\textbf{Conflict cases.} Two concurrent differing modifications of one structure
produce a \texttt{StructuralFieldCollision} on the field \texttt{cross\_cutting}.

\textbf{Undo semantics.} Value-restoring (Section~\ref{sec:k0:undo}): undo
restores the structure's chain-predecessor value unless superseded.

\textbf{Re-anchoring.} If the target structure is tombstoned, the modification is
a no-op; re-deriving the endpoints lets a subsequent endpoint tombstone re-anchor
the structure through the standard table (see DeleteEvent).

\section{ChangeRegionTimeModel}
\label{sec:k0:change-region-time-model}

\textbf{Payload schema.} \texttt{ChangeRegionTimeModelOp \{ region: RegionId,
new\_time\_model: RegionTimeModel, declared\_incompatible: Vec<EventId>,
remapping: PositionRemapping \}} --- the full target model value (v1).

\textbf{Canonical encoding.} \texttt{region}, the length-framed
\texttt{new\_time\_model} value, the canonically-ordered
\texttt{declared\_incompatible} set, then \texttt{remapping}.

\textbf{Reduction rule.} Structural migration. The region adopts the target time
model. Graph-aware reduction derives coordinate-kind incompatibilities from the
region's events (and from the remapping coverage) and refuses a migration that
would violate the coordinate discipline.

\textbf{Conflict cases.} Concurrent same-region migrations produce a
\texttt{StructuralFieldCollision} on the field \texttt{time\_model}; a migration
with incompatible events produces a \texttt{TimeModelMigrationFailure} naming the
region and the incompatible events. A causally-later migration is re-evaluated
against the first migration's graph rather than conflicting.

\textbf{Undo semantics.} Undo restores the region's prior time model under the
active policy.

\textbf{Re-anchoring.} Not applicable.

\begin{openquestion}
\textbf{P11-C6.} The rich migration payload --- a coordinate converter rather than a
\texttt{declared\_incompatible} list plus a \texttt{PositionRemapping} --- remains
the catalog's to design when graph-aware migration is the only reduction path.
\end{openquestion}

\section{Structural Containers}
\label{sec:k0:structural-containers}

\textbf{Payload schema.} Three create/delete pairs over the region hierarchy:
\texttt{CreateRegionOp \{ region: Region \}} /
\texttt{DeleteRegionOp \{ region: RegionId \}};
\texttt{CreateStaffInstanceOp \{ region: RegionId, instance: StaffInstance \}} /
\texttt{DeleteStaffInstanceOp \{ staff\_instance: StaffInstanceId \}};
\texttt{CreateVoiceOp \{ staff\_instance: StaffInstanceId, voice: Voice \}} /
\texttt{DeleteVoiceOp \{ voice: VoiceId \}}. Each create carries the full
container value (v1); the reduction preconditions it bears \emph{no typed child
object} --- an empty container.

\textbf{Canonical encoding.} Create: the parent id (where the schema names one),
then the length-framed canonical bytes of the container value. Delete: the
container id.

\textbf{Reduction rule.} Set-union creation of an \emph{empty} container, and an
\emph{empty-only} delete-wins tombstone. A create mints the container live if its
id is fresh and (for staff instance and voice) its parent is live; it
preconditions the carried value to bear no typed child object (a region: no staff
instances, barline-alignment groups, or graphic objects; a staff instance: no
voices or measures; a voice: no events), since those carry distinct
\texttt{TypedObjectId}s the reducer mints separately --- so contents are added by
subsequent operations. A delete is a delete-wins tombstone, but a
\emph{precondition no-op} (\texttt{ContainerNotEmpty}) unless the container has no
live children --- the caller deletes contents first. Graph-aware reduction adds
or removes the container and maintains the region's staff extent so
\texttt{RegionExtents} stays satisfied.

\textbf{Conflict cases.} None at reduction time: creation is set-union (a repeat
create is idempotent), and the empty-only delete is a deterministic precondition
gate, not a conflict.

\textbf{Undo semantics.} Undo of a \emph{create} tombstones the minted container
(Section~\ref{sec:k0:undo}); \texttt{StrictInverse} conflicts if it was
concurrently mutated, with the policy treatment as for InsertEvent. A
\emph{delete} mints nothing, so the prototype's minted-object undo does not
re-introduce it (P11-C8).

\textbf{Re-anchoring.} Not applicable (the containers are minted/tombstoned by id;
the empty-only precondition means a delete never strands live children).

\section{CreateStaff}
\label{sec:k0:create-staff}

\textbf{Payload schema.} \texttt{CreateStaffOp \{ staff: Staff \}} --- the full
global-staff value (v1): identity, name, abbreviation, instrument reference,
default staff-line configuration, and optional group membership.

\textbf{Canonical encoding.} The length-framed canonical bytes of
\texttt{staff}.

\textbf{Reduction rule.} Set-union creation of a global \texttt{Staff} on the
score root, completing the structural-container family
(Section~\ref{sec:k0:structural-containers}) upward: staff \emph{instances}
reference global staves, and until this primitive existed a resolvable staff
could only be base-seeded. A create mints the staff live if its id is fresh; a
repeat create carrying a byte-identical value reduces idempotently, and a
create whose id is already live with a \emph{differing} value is a
precondition no-op with the appended reason
\texttt{RecreateContentMismatch} (discriminant 13; ratified Pass~12, closing
P12-K9 --- the former \texttt{TargetMissing} reuse misnamed the situation:
the target is not missing, its content disagrees). The same reason applies
to the carried \texttt{TimeSignature} (Section~\ref{sec:k0:meter-tempo}),
the other re-create site at which the reducer retains the carried value for
comparison. The structural-container creates
(Section~\ref{sec:k0:structural-containers}) are plain set-union --- any
repeat create of a live id reads \texttt{AlreadyApplied} without value
comparison, since the carried value is preconditioned empty of children. Graph-aware reduction additionally preconditions that the
referenced instrument is live and, when \texttt{group} is present, that the
staff group resolves --- the mint must leave the graph satisfying the
reference-resolution invariants.

With staves mintable, \texttt{CreateStaffInstance}
(Section~\ref{sec:k0:structural-containers}) preconditions that the
instance's referenced \texttt{Staff} is live (previously the reference was
satisfiable only from the seeded base, so the check was vacuous).

\textbf{Conflict cases.} None at reduction time (set-union; the differing-value
re-create is a precondition gate, not a conflict).

\textbf{Undo semantics.} Undo of a create tombstones the minted staff
(Section~\ref{sec:k0:undo}); \texttt{StrictInverse} conflicts if a live staff
instance references it (tombstoning it would strand the instance).

\textbf{Re-anchoring.} Not applicable (a staff mint references no tombstonable
anchor; there is no \texttt{DeleteStaff} in this catalogue revision --- an
empty-only staff delete mirroring the container discipline is a later
schema-fill).

\section{Repeat Structures}
\label{sec:k0:repeat-structures}

Ratified with the schema-major-2 revision: the dedicated authoring pair for
\texttt{RepeatStructure} (core specification Chapter~5
\sectionsc{Repeat Structures}). Repeats live in the cross-cutting registry
but are \emph{not} \texttt{CrossCuttingValue} kinds on the wire --- the
cross-cutting operations admit only
\texttt{Tie}/\texttt{Slur}/\texttt{Beam}/\texttt{Spanner}
(Section~\ref{sec:k0:create-cross-cutting}) --- so the pair is a
first-class primitive mirroring the cross-cutting disciplines.

\textbf{Payload schema.} \texttt{CreateRepeatStructureOp \{ repeat:
RepeatStructure \}} --- the full value (schema major 2): identity, the
\texttt{start}/\texttt{end} anchors, the \texttt{RepeatKind}, and the
volta list. \texttt{DeleteRepeatStructureOp \{ repeat: RepeatStructureId
\}}.

\textbf{Canonical encoding.} Create: the length-framed canonical bytes of
\texttt{repeat} (the schema-major-2 layout). Delete: the bare identifier.
The create's payload embeds the v2 \texttt{RepeatStructure} layout
unconditionally (\texttt{kind} and \texttt{voltas} are not optional
fields), so under minimal stamping a block carrying one stamps schema
major~2 --- the create is \emph{born at v2}. The delete's payload is an
identifier --- a major-0 layout --- so minimal stamping gives its blocks
major~0; the kind discriminant itself is an append-only schema-\emph{minor}
vocabulary event (the same append mechanism as the Phase-3 tranche; the
stamp itself always follows minimal stamping over the payload, per the
Binary Format companion \sectionsc{Schema Major 2}).

\textbf{Reduction rule.} Set-union creation and a delete-wins tombstone,
mirroring the cross-cutting family. A create mints the repeat live if its
id is fresh and \emph{every} event-referencing anchor site resolves to a
live event --- \texttt{start}/\texttt{end}, the kind's jump targets
(\texttt{DaCapo.end\_target},
\texttt{DalSegno.segno}/\texttt{end\_target}), and each volta's
\texttt{start}/\texttt{end} --- since the mint must leave the graph
satisfying the reference-resolution invariants; a dead anchor is a
precondition no-op (\texttt{TargetMissing}). A repeat create of a live id
reads \texttt{AlreadyApplied} without value comparison (the cross-cutting
discipline; the \texttt{RecreateContentMismatch} scope of
Section~\ref{sec:k0:create-staff} is unchanged). A delete tombstones the
repeat (idempotent on re-delete, delete-wins); deleting a missing id is a
no-op precondition failure. Graph-aware reduction adds or removes the
structure from the cross-cutting registry.

\textbf{Conflict cases.} None at reduction time (set-union creation;
delete-wins is idempotent).

\textbf{Undo semantics.} Undo of a create tombstones the minted repeat
(Section~\ref{sec:k0:undo}). A delete mints nothing, so the prototype's
minted-object undo does not re-introduce the tombstoned structure
(P11-C8).

\textbf{Re-anchoring.} The structure participates in the re-anchoring rule
table when a referenced event is later tombstoned: re-anchor to the
nearest surviving anchor, cascade-delete only when none survives --- as
for spanners, spanning \emph{every} anchor site
(\texttt{start}/\texttt{end}, jump targets, volta spans). Among multiple
surviving candidates --- a case slurs and spanners never present, since
their sole other endpoint is the forced survivor --- ``nearest'' is
currently realized as the deterministic identifier-order minimum (the
same tie-break the two-endpoint collapse already used); proximity-aware
(four-key) selection over a repeat's surviving sites is a deferred
refinement, exactly as the spanner row defers per-kind proximity bounds.
The rule row is the core specification's (Chapter~6 re-anchoring rule
table, ratified with this pair).

\textbf{Authoring advisory.} The volta well-formedness constraints of
core Chapter~5 (endings non-empty, 1-based, \emph{strictly} ascending) are
\emph{advisory}: surfaced at authoring time under interactive validation,
never enforced under reduction.

\section{SetUserSystemBreak}
\label{sec:k0:set-user-system-break}

\textbf{Payload schema.} \texttt{SetUserSystemBreakOp \{ region: RegionId,
anchor: TimeAnchor, present: bool \}} --- the full anchor value (v1).

\textbf{Canonical encoding.} \texttt{region}, the length-framed \texttt{anchor}
value, then the boolean.

\textbf{Reduction rule.} A last-writer-wins advisory. The break preference is
recorded for the region keyed by the anchor's \emph{resolved musical position};
graph-aware reduction adds or removes the anchor from the region's user
system-break list.

\textbf{Conflict cases.} None (LWW advisory).

\textbf{Undo semantics.} Undo restores the prior advisory value for the
\texttt{(region, resolved-position)} key.

\textbf{Re-anchoring.} Not applicable in the prototype (the advisory is keyed by
resolved position; a tombstoned anchor target degrades to the region origin).

\section{Score Settings}
\label{sec:k0:score-settings}

\textbf{Payload schema.} Three score-level field overwrites:
\texttt{SetMetadataOp \{ metadata: ScoreMetadata \}} overwrites the score
singleton; \texttt{SetMetricGridOp \{ region: RegionId, grid: Option<MetricGrid>
\}} overwrites (or clears) a region's default metric grid;
\texttt{SetUserPageBreakOp \{ region: RegionId, anchor: TimeAnchor, present: bool
\}} is the page-break sibling of SetUserSystemBreak.

\textbf{Canonical encoding.} Metadata: the length-framed \texttt{metadata} value.
Metric grid: \texttt{region}, then an \texttt{Option} discriminant and (when
present) the length-framed \texttt{grid} value. Page break: \texttt{region}, the
length-framed \texttt{anchor} value, then the boolean.

\textbf{Reduction rule.} Three field overwrites differing only in discipline.
\emph{SetMetadata} is an \textbf{advisory} last-writer-wins: the latest write in
canonical order silently wins and the operation always applies --- no working
state and no conflict (the same discipline as SetUserSystemBreak, on the score
singleton). \emph{SetMetricGrid} is a \textbf{structural} field overwrite keyed
by region: precondition the region is live and staff-based (a FreeGraphic region
has no metric-grid slot --- the op is a no-op there), and reject a grid whose
meter sequence names a time signature that is not live (the Chapter~5 invariant
forbids installing such a grid). \emph{SetUserPageBreak} is a canonical LWW
advisory keyed by the anchor's resolved musical position, with the same
staff-based precondition. Graph-aware reduction overwrites the metadata singleton,
sets the region's default metric grid, or adds/removes the page-break anchor
under its resolved-position key (so two anchors resolving to one position occupy a
single slot).

\textbf{Conflict cases.} SetMetadata and SetUserPageBreak: none (advisory LWW).
SetMetricGrid: two concurrent differing grids for one region produce a
\texttt{StructuralFieldCollision} on the field \texttt{metric\_grid}.

\textbf{Undo semantics.} All three are value-restoring field overwrites
(Section~\ref{sec:k0:undo}): undo restores the prior metadata, grid, or break
preference from the key's write chain unless superseded.

\textbf{Re-anchoring.} The advisory breaks degrade as for SetUserSystemBreak; the
metric grid and metadata are keyed by region / singleton and do not re-anchor (a
deleted region's settings are no-ops --- \texttt{TargetMissing}).

\section{Meter and Tempo Overwrites}
\label{sec:k0:meter-tempo}

\textbf{Payload schema.} The finer-grained metric-model overwrites beneath the
whole-grid \texttt{SetMetricGrid} (Section~\ref{sec:k0:score-settings}):
\texttt{SetTimeSignatureOp \{ region: RegionId, anchor: TimeAnchor,
time\_signature: Option<TimeSignature> \}} sets, replaces, or (\texttt{None})
removes the single \texttt{MeterChange} at the anchor's resolved musical
position in the region's default metric grid, carrying the full
\texttt{TimeSignature} value (v1); \texttt{SetTempoSegmentOp \{ region:
Option<RegionId>, start: TimeAnchor, segment: Option<TempoSegment> \}} sets,
replaces, or removes the single tempo segment starting at the resolved
position, in the score-level tempo map (\texttt{region: None}) or the region's
local map (\texttt{Some}; a set on a region with no local map creates one).

\textbf{Canonical encoding.} Time signature: \texttt{region}, the
length-framed \texttt{anchor}, then an \texttt{Option} discriminant and (when
present) the length-framed \texttt{time\_signature} value. Tempo segment: an
\texttt{Option} discriminant and (when present) \texttt{region}, then the
length-framed \texttt{start}, then an \texttt{Option} discriminant and (when
present) the length-framed \texttt{segment}.

\textbf{Reduction rule.} Last-writer-wins structural overwrites keyed by
\texttt{(region, resolved position)} (time signature) and \texttt{(scope,
resolved start)} (tempo segment). A carried \texttt{TimeSignature} is minted
set-union under the same discipline as \texttt{CreateStaff}: fresh id mints;
byte-identical re-carry is idempotent; a differing value under a live id is a
precondition no-op (\texttt{RecreateContentMismatch},
Section~\ref{sec:k0:create-staff}). The time-signature value's beat-group sum is validated at
construction and again at decode, so a malformed value never reaches
reduction. A tempo-segment write preconditions that the \emph{resulting} map
is well-formed (segments ordered and non-overlapping; a non-constant shape
carries its end data; the carried segment's own start equals the operation's
\texttt{start} key) --- a write that would malform the map is refused as a
precondition no-op (\texttt{TempoMapMalformed}). Graph-aware reduction applies
the meter change to \texttt{default\_metric\_grid.meter\_sequence} and the
segment to the scoped tempo map.

\textbf{Conflict cases.} Two concurrent differing writes of one key produce a
\texttt{StructuralFieldCollision} on the field \texttt{meter\_sequence} or
\texttt{tempo\_segments} respectively, with the standard winner/loser
recording; identical concurrent writes reduce idempotently.

\textbf{Undo semantics.} Value-restoring per Section~\ref{sec:k0:undo}: undo
restores the key's chain-predecessor value (or its absence).

\textbf{Re-anchoring.} An event-anchored \texttt{anchor}/\texttt{start} whose
event is later tombstoned degrades by the framework's anchor rules; the
overwrite keys on the \emph{resolved} position, so the recorded change
survives its anchor.

\begin{openquestion}
\textbf{P12-C5.} A mid-region meter change authored by
\texttt{SetTimeSignature} reduces cleanly and materialises into the grid, but
the notational-decomposition pre-pass currently honours only a region's
\emph{first} governing meter (P12-H4's single-meter simplification), so the
derived notation ignores the change until multi-meter decomposition lands.
The reduction-level semantics are pinned here; the derived-annotation gap is
P12-H4's.
\end{openquestion}

\section{SetStaffLayout}
\label{sec:k0:set-staff-layout}

\textbf{Payload schema.} \texttt{SetStaffLayoutOp \{ staff\_instance:
StaffInstanceId, instrument\_override: Option<InstrumentId>,
staff\_lines\_override: Option<StaffLineConfiguration>, visible: bool \}} ---
the non-break layout advisories with a graph home: the staff instance's three
inline advisory fields, overwritten as a unit.

\textbf{Canonical encoding.} \texttt{staff\_instance}, an \texttt{Option}
discriminant and (when present) \texttt{instrument\_override}, an
\texttt{Option} discriminant and (when present) the length-framed
\texttt{staff\_lines\_override}, then the boolean.

\textbf{Reduction rule.} A last-writer-wins \emph{advisory} overwrite keyed by
\texttt{staff\_instance}. Preconditions: the staff instance is live; a present
\texttt{instrument\_override} resolves to a live instrument under graph-aware
reduction. The richer engraving-override vocabulary (stem direction, notehead
shape, custom positions) has no durable graph home yet and remains projected
layout state --- extending this primitive to cover it is staged with the
data-model expansion.

\textbf{Conflict cases.} None (LWW advisory).

\textbf{Undo semantics.} Value-restoring per Section~\ref{sec:k0:undo}.

\textbf{Re-anchoring.} If the staff instance is tombstoned, the overwrite is a
no-op (\texttt{TargetTombstoned}).

\section{DeclareTransaction}
\label{sec:k0:declare-transaction}

\textbf{Payload schema.} \texttt{TransactionDescriptor \{ id: TransactionId,
label: String, category: Option<TransactionCategory> \}}. Value-complete in v0
and unchanged.

\textbf{Reduction rule.} Records the descriptor. Member primitives reference the
transaction id and \MUST{} causally depend on the descriptor; the members reduce
atomically (all-or-nothing) in canonical order.

\textbf{Conflict cases.} A missing descriptor or a member that does not causally
follow it produces a \texttt{TransactionConflict}; any member failure rolls back
the whole transaction and all members read \texttt{NoOp\{TransactionConflict\}}.

\textbf{Undo / re-anchoring.} Transactions are the unit of undo
(Section~\ref{sec:k0:undo}); re-anchoring is per member.

\section{ResolveConflict (meta-operation)}
\label{sec:k0:resolve-conflict}

\textbf{Payload schema.} \texttt{ResolveConflictPayload \{ target: ConflictId,
action: ResolutionAction \}}. Value-complete.

\textbf{Reduction rule.} Transitions the target conflict's resolution state. An
action of \texttt{Dismiss} reaches the \texttt{Dismissed} state; any other action
reaches \texttt{Resolved}. Re-resolving with the same action is idempotent; two
concurrent resolves with differing actions produce a meta-conflict.

\emph{Beyond the concurrent case (ratified Pass~12, closing P12-K4).} The
earliest-applied-resolve-governs rule is \emph{universal}: a causally-later
resolve with a differing action does not supersede the first --- it reduces
\texttt{AlreadyApplied} --- and \emph{any} resolve targeting a
\texttt{Dismissed} conflict likewise reads \texttt{AlreadyApplied}.
Intentional re-resolution is deliberately outside the v1 operation set; a
future dedicated operation (a \texttt{ReopenConflict}-class primitive) is the
sanctioned path if it is ever needed. The meta-conflict record names both
resolver operation ids in \texttt{caused\_by}; conflict records themselves
have no \texttt{TypedObjectId} kind (deliberate --- they are materialized
state, not graph objects), so the contested conflict is identified by the
\texttt{equivocation}-style field key, not an object reference.

\begin{rationale}
Pass~11 added \texttt{ResolutionAction::Dismiss} (item 2.5) precisely so the
\texttt{Dismissed} state is reachable by an authored operation rather than merely
representable. The catalog records that \texttt{Dismiss} is the action that
selects it (resolving the v0 ambiguity P11-C10).
\end{rationale}

\section{ResolveEquivocation (meta-operation)}
\label{sec:k0:resolve-equivocation}

\textbf{Payload schema.} \texttt{ResolveEquivocationPayload \{ target:
OperationId, chosen: EnvelopeHash \}} --- the equivocated slot and the
candidate envelope (by canonical-bytes hash) that shall stand. Value-complete.

\textbf{Canonical encoding.} \texttt{target} (16 canonical bytes), then
\texttt{chosen} (32 bytes), per the codec baseline.

\textbf{Reduction rule.} Order-independent promotion of an equivocated slot
(core specification Chapter~6, \sectionsc{Equivocation}): when the operation
set holds an \texttt{Equivocated} slot for \texttt{target} and \texttt{chosen}
names one of its candidates, the slot reduces as if it had always been
\texttt{Single} with the chosen envelope --- the chosen candidate contributes
to canonical reduction at its own canonical position, and operations that were
pending on the equivocated id unblock. Among multiple resolves naming the same
slot, the one earliest in canonical order governs; a later resolve naming the
\emph{same} candidate reduces idempotently (\texttt{AlreadyApplied}). The
resolve operation must itself occupy a \texttt{Single} slot; an equivocated
resolve is excluded from reduction like any other equivocated slot. A resolved
slot records no \texttt{OperationSlotEquivocated} anomaly; the losing
candidates remain in the diagnostic candidate store only.

\textbf{Conflict cases.} Two resolves of one slot naming \emph{differing}
candidates produce a \texttt{StructuralFieldCollision} meta-conflict on the
field \texttt{equivocation\_resolution}, recording the governing resolve
(earlier in canonical order) as winner and the later as loser, with both
operations in \texttt{caused\_by} --- the same discipline as
\texttt{ResolveConflict} meta-conflicts. Preconditions: a resolve whose
\texttt{target} is not an equivocated slot, or whose \texttt{chosen} is not
among the slot's candidates, is a precondition no-op.

\textbf{Edge semantics (ratified Pass~12, closing P12-K6).} Promotion is
\emph{single-pass}, not fixpoint: a promoted candidate that is itself a
\texttt{ResolveEquivocation} does not govern a further promotion in the same
reduction. A resolve in a \emph{quarantined} replica segment is excluded from
reduction and never governs. A resolve held \emph{pending} by its own causal
gaps still governs promotion --- the verdict is a pure function of the slot
map (set-level), while the resolve's own effect stays pending. The
invalid-target/invalid-chosen no-op reuses \texttt{TargetMissing}; a dedicated
reason was considered and rejected (the appended-reason budget is spent where
a distinct verdict changes caller behavior, which it does not here).

\textbf{Undo semantics.} Mints nothing; not inverted under the prototype's
minted-object undo (P11-C8).

\textbf{Re-anchoring.} Not applicable (the payload references an operation
slot, not a graph object).

\begin{rationale}
The core specification names three resolution paths for an equivocated slot:
transport-level reconciliation, this explicit operation, and a
profile-declared deterministic selection policy. This entry pins the schema
for the explicit-operation path, which the core specification previously named
only in prose. The profile-policy path is \emph{deferred with a named landing
site} (Pass~12 disposition of P12-K5): v1 profiles declare \emph{no}
selection function, and the hook's definition belongs to the Profile
Conformance companion when it is written --- the reducer deliberately carries
no policy hook until a profile can declare one.
\end{rationale}

\section{UndoTransaction (meta-operation)}
\label{sec:k0:undo}

\textbf{Payload schema.} \texttt{UndoTransactionPayload \{ target: TransactionId,
policy: UndoPolicy \}}, with \texttt{UndoPolicy} one of \texttt{StrictInverse},
\texttt{BestEffort}, \texttt{Cascade}. Value-complete.

\textbf{Reduction rule.} A forward compensating edit computed against the
materialised state at the undo's canonical position (never literal time
travel). The compensation has two parts.

\emph{Minted-object tombstoning} (as before): every object the target
transaction minted is tombstoned. \texttt{StrictInverse} conflicts
(\texttt{TombstonedTarget}) if any minted object was already tombstoned;
\texttt{BestEffort} tombstones the survivors.

\emph{Value restoration} (this revision): for every last-writer-wins overwrite
the target transaction performed --- event and identified-pitch modification,
respelling, cross-cutting modification, metadata, metric grid, meter change,
tempo segment, staff layout, and the user break advisories --- the reducer
maintains, per overwritten key, the \emph{canonical-order write chain} of
(writer, value) pairs. Undoing the transaction restores each written key to
its chain-predecessor value (or its absence, where the transaction introduced
the first value), \emph{provided the transaction's write is still the key's
last writer}. Because the chain is keyed by canonical order, the restored
value is a pure function of the operation set: permutation-invariant by
construction. When a causally-later or canonically-later write has
superseded the key, \texttt{StrictInverse} refuses the whole undo with a
\texttt{TransactionConflict} conflict naming the undo and the superseding
writer; \texttt{BestEffort} restores the still-last-written keys and skips
the superseded ones. Restorations are expressed in the effect status (a fully
clean compensation is \texttt{Applied}; a mixed one is
\texttt{AppliedWithRepair} carrying only the tombstone repairs) --- no new
repair vocabulary.

\emph{Strand-blocks (ratified Pass~12, closing P12-K10).} A
\texttt{StrictInverse} undo that refuses to tombstone a minted object still
referenced by a live non-member (e.g., a staff whose instance survives outside
the transaction) records \texttt{ConflictKind::TransactionConflict} --- the
reuse is blessed: the strand-block \emph{is} a transaction-scoped conflict of
the undo, and the conflict record's affected objects and description carry the
strand detail. No dedicated undo conflict kind is added.

\emph{Idempotence asymmetry (ratified Pass~12, closing P12-K11).} An undo's
value restorations enter the write chains as ordinary writes by the undo
operation --- no distinguished undo provenance. Consequently a \emph{second}
undo of the same transaction finds each restored key superseded by the first
undo and refuses (\texttt{Conflicted} under \texttt{StrictInverse}, skipped
under \texttt{BestEffort}), while \emph{absence} restorations (not
representable as chain writes) repeat idempotently. The asymmetry is
normative, documented behavior; a chain-native undo provenance would be
revisited only under the deferred undo-as-operation (streaming-consistent
undo) design, which subsumes this question.

\emph{Still deferred} (P11-C8, narrowed): re-introducing content tombstoned by
\emph{delete} primitives (a deterministic resurrection needs a system-derived
identifier derivation the ratified closed tag set does not yet include); and
\texttt{Cascade}'s dependent-closure computation --- \texttt{Cascade} remains
\texttt{StrictInverse} over the same set. \texttt{Transpose} inversion is no
longer listed: \texttt{TransposeInterval} is undone by value restoration, not by
applying an inverse interval (Push~4a, closing P12-K2). An inverse
$(-d, -c)$ does exist wherever both components are negatable --- the reduction
never saturates --- but $\texttt{i32::MIN}$ has none, so the inverse is partial
and undo does not depend on it. The frozen \texttt{Transpose} is undone by
neither.

% ===========================================================================
\chapter{v0 \texorpdfstring{$\rightarrow$}{->} v1 Payload Migration}
\label{ch:migration}

A v0 envelope carries an identifier-only payload; a v1 envelope carries the
value-typed payload this catalog defines. The two forms do not coexist as
permanent dialects (that would double the reducer surface forever); instead the
catalog ships a \textbf{one-time migration} that lifts a v0 envelope to v1 using
the score graph as context, applied once on read. Production code carries only v1
payloads; v0 envelopes survive only as a regression corpus.

\begin{requirement}
\label{req:migration:properties}
The migration \texttt{migrate\_v0\_envelope(v0, context: \&Score)} \MUST{} be
\textbf{deterministic} (two implementations migrating the same v0 envelope
against the same context produce byte-identical v1 envelopes) and
\textbf{equivalence-preserving} (a v0 envelope and its v1 migration reduce to
byte-identical canonical \texttt{MaterializedState}). When a value cannot be
reconstructed from the v0 projection plus the context, the migration \MUST{}
report the envelope unmigratable rather than fabricate a value, and the bundle
opens read-only.
\end{requirement}

The reference implementation (\texttt{epiphany-ops::migrate}) reconstructs the
\texttt{InsertEvent} event, the \texttt{DeleteEvent} compensation, the
\texttt{ChangeRegionTimeModel} model, the \texttt{SetUserSystemBreak} anchor, and
the event-anchored cross-cutting structures (\texttt{Tie} / \texttt{Slur} /
\texttt{Beam}) self-containedly from the v0 projection; a \texttt{Spanner},
anchored by \texttt{TimeAnchor}s rather than event endpoints, remains unmigratable
until the projection carries them. It recovers a \texttt{RespellPitch} spelling
from the context (P12-K1,
Section~\ref{sec:k0:respell-pitch}). The migration's merge gate
(\texttt{epiphany-testkit::migration}) drives the inverse direction --- projecting
a v1 corpus to v0 and migrating it back --- and asserts byte-identical reduction
plus a non-vacuity guard.

% ===========================================================================
\chapter{K1 --- Framework Slots (Phase 3)}
\label{ch:k1}

This chapter drafted the remaining catalogue items as framework slots. The
Phase-2 \textbf{M2} expansion (the broad-K0 groups in \texttt{epiphany-ops})
implemented four groups of them; with the \textbf{M2e} catalogue expansion their
full per-primitive schemas now appear in Chapter~\ref{ch:k0}, so they are
\emph{normative under the Phase-2 profile} and an implementation \MUST{}
\emph{not} reject them. They are cross-referenced first. The genuinely Phase-3
slots that remain \textbf{unavailable} are listed second: an implementation
\MUST{} reject an operation of one of \emph{those} kinds. Each remaining slot is a
schema-fill against the template of Chapter~\ref{ch:framework}; adding one is not
a fresh design.

\section*{Implemented since M2 (now in Chapter~\ref{ch:k0})}

\begin{description}
  \item[Modify event; identified-pitch operations; transpose]
    M2 Group~1 --- Sections~\ref{sec:k0:modify-event},
    \ref{sec:k0:identified-pitch}, and \ref{sec:k0:transpose}.
  \item[Delete / modify cross-cutting]
    M2 Group~2 --- Sections~\ref{sec:k0:delete-cross-cutting} and
    \ref{sec:k0:modify-cross-cutting} (creation is
    Section~\ref{sec:k0:create-cross-cutting}).
  \item[Create / delete region / staff instance / voice]
    M2 Group~3 --- Section~\ref{sec:k0:structural-containers} (set-union creation
    and the empty-only delete).
  \item[Set metadata / metric grid / user page break]
    M2 Group~4 --- Section~\ref{sec:k0:score-settings} (advisory metadata,
    structural metric grid, advisory page break).
\end{description}

\section*{Implemented in the Phase-3 first tranche (now in Chapter~\ref{ch:k0})}

\begin{description}
  \item[Create staff; set time signature / tempo segment; set layout]
    Sections~\ref{sec:k0:create-staff}, \ref{sec:k0:meter-tempo}, and
    \ref{sec:k0:set-staff-layout}. The disciplines are as drafted: set-union
    creation, LWW structural overwrite, and LWW advisory respectively.
\end{description}

\section*{Added in the schema-major-2 revision (Chapter~\ref{ch:k0})}

\begin{description}
  \item[Create / delete repeat structure]
    Section~\ref{sec:k0:repeat-structures} --- net-new primitives (never
    drafted as framework slots). Set-union creation with the
    all-anchors-live precondition and a delete-wins tombstone; the create
    is born at v2, the delete stamps major~0.
\end{description}

\section*{Retired slots (ratified Pass~12: outside the operation set)}

\begin{description}
  \item[Create score / canvas --- retired, closing P12-K8]
    The document root and the canvas are \emph{structural givens}, not
    operation products: \texttt{TypedObjectId} has no Canvas kind, the root is
    never op-minted, and genesis is normatively the empty-document constructor
    plus bundle creation, outside the operation set (core specification,
    Chapter~5 \sectionsc{The Canvas}). These are not ``unavailable slots''
    awaiting a design --- no operation kind will be assigned to them. The
    decision is revisited only if an addressable multi-canvas model is adopted
    at a future schema major.
\end{description}

\begin{nongoal}
The full $60$--$80$-primitive catalogue is not a Phase-2 deliverable. The
framework (Chapter~\ref{ch:framework}) and the K0 representative set
(Chapter~\ref{ch:k0}) are sufficient to exercise every reduction discipline; the
remaining primitives are Phase-3 schema-fill.
\end{nongoal}

\end{document}
