% !TEX program = xelatex
%
% Epiphany --- Text Projection (companion specification)
% Companion to the Core Specification. Compile with XeLaTeX.
%
% This document is versioned independently of the Core Specification
% (independent semver; see the Versioning note in the front matter). Its preamble
% is intentionally a self-contained copy of the core specification's preamble so
% the two documents build independently; factoring a shared preamble file is a
% later cleanup, not a v0.1 deliverable.

\documentclass[11pt,letterpaper]{report}

% ---------------------------------------------------------------------------
% Packages
% ---------------------------------------------------------------------------
\usepackage{fontspec}
\usepackage{geometry}
\geometry{
  letterpaper,
  top=1.05in,
  bottom=1.05in,
  left=1.15in,
  right=1.15in,
  headheight=15pt
}

\usepackage[english]{babel}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tcolorbox}
\tcbuselibrary{breakable, skins}

% ---------------------------------------------------------------------------
% Color palette (shared with the core specification)
% ---------------------------------------------------------------------------
\definecolor{epiphanyteal}{HTML}{1A4044}
\definecolor{epiphanygold}{HTML}{8E6E2E}
\definecolor{epiphanyink}{HTML}{1F1B16}
\definecolor{epiphanyslate}{HTML}{6B6660}
\definecolor{epiphanycream}{HTML}{F8F4ED}
\definecolor{epiphanymist}{HTML}{ECE8E0}
\definecolor{epiphanycode}{HTML}{2A2520}
\definecolor{epiphanycrimson}{HTML}{7A2424}

\hypersetup{
  colorlinks=true,
  linkcolor=epiphanyteal,
  citecolor=epiphanyteal,
  urlcolor=epiphanygold,
  pdftitle={Epiphany --- Text Projection},
  pdfauthor={The Epiphany Project},
  pdfsubject={Text Projection companion for the Epiphany music notation platform},
  pdfkeywords={music notation, operations, CRDT, reduction, serialization},
  bookmarksnumbered=true,
  bookmarksopen=true
}

% ---------------------------------------------------------------------------
% Typography (shared with the core specification)
% ---------------------------------------------------------------------------
\setmainfont{TeX Gyre Pagella}[Numbers={OldStyle, Proportional}, Ligatures={TeX, Common}]
\setsansfont{TeX Gyre Heros}[Scale=0.94, Ligatures={TeX, Common}]
% No TeX ligatures in the mono font: `tlig` maps " to a right curly quote and --
% to an en dash. This document's grammar quotes terminals with U+0022 and spells
% escapes as backslash sequences, so a substituted glyph would misstate the syntax.
\setmonofont{TeX Gyre Cursor}[Scale=0.88]
\newfontfamily\titlefont{TeX Gyre Pagella}[Numbers={OldStyle}, Ligatures={TeX, Common}]
\newcommand{\tablenums}[1]{{\addfontfeatures{Numbers={Lining,Tabular}}#1}}
\newcommand{\sectionsc}[1]{{\addfontfeatures{Letters=SmallCaps}#1}}

% ---------------------------------------------------------------------------
% Section styling (shared with the core specification)
% ---------------------------------------------------------------------------
\titleformat{\chapter}[display]
  {\normalfont\filright}
  {\raggedright\color{epiphanygold}\fontsize{14pt}{16pt}\selectfont
    \scshape Chapter\ \thechapter}
  {16pt}
  {\raggedright\color{epiphanyteal}\fontsize{32pt}{36pt}\selectfont\bfseries}
  [\vspace{4pt}{\color{epiphanygold}\rule{2in}{0.6pt}}]
\titlespacing*{\chapter}{0pt}{-20pt}{30pt}
\titleformat{\section}
  {\normalfont\Large\bfseries\color{epiphanyteal}}
  {\color{epiphanygold}\thesection}{1em}{}
\titleformat{\subsection}
  {\normalfont\large\bfseries\color{epiphanyteal}}
  {\color{epiphanygold}\thesubsection}{1em}{}
\titleformat{\subsubsection}
  {\normalfont\normalsize\bfseries\color{epiphanyink}}
  {\thesubsubsection}{1em}{}

% ---------------------------------------------------------------------------
% Headers and footers (shared with the core specification)
% ---------------------------------------------------------------------------
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\fancyhead[L]{\small\scshape\color{epiphanyslate}Epiphany --- Text Projection}
\fancyhead[R]{\small\itshape\color{epiphanyslate}\leftmark}
\fancyfoot[C]{\small\color{epiphanyslate}\thepage}
\renewcommand{\headrule}{
  \color{epiphanygold!50}\hrule width\headwidth height 0.4pt
  \vspace{1pt}
  \color{epiphanygold!30}\hrule width\headwidth height 0.2pt
}

% ---------------------------------------------------------------------------
% Code listing style (shared with the core specification)
% ---------------------------------------------------------------------------
\lstdefinelanguage{Rust}{
  keywords={fn,let,mut,pub,struct,enum,impl,trait,for,in,if,else,match,return,
           use,mod,crate,self,Self,as,where,move,async,await,const,static,
           ref,type,unsafe,extern,dyn,box,break,continue,loop,while},
  keywordstyle=\color{epiphanyteal}\bfseries,
  ndkeywords={i8,i16,i32,i64,i128,u8,u16,u32,u64,u128,f32,f64,bool,char,str,
              String,Vec,Option,Result,Box,Rc,Arc,HashMap,BTreeMap,
              NonZeroU16,NonZeroU32,NonZeroU64,Duration,Timestamp},
  ndkeywordstyle=\color{epiphanygold}\bfseries,
  sensitive=true,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{epiphanyslate}\itshape,
  stringstyle=\color{epiphanycrimson},
  morestring=[b]",
  morestring=[b]'
}
\lstset{
  basicstyle=\ttfamily\small\color{epiphanycode},
  backgroundcolor=\color{epiphanycream},
  frame=leftline,
  rulecolor=\color{epiphanygold!60},
  framesep=8pt,
  framerule=1.5pt,
  xleftmargin=10pt,
  xrightmargin=4pt,
  breaklines=true,
  showstringspaces=false,
  numberstyle=\tiny\color{epiphanyslate},
  numbersep=10pt,
  captionpos=b,
  aboveskip=10pt,
  belowskip=10pt,
  language=Rust
}

% ---------------------------------------------------------------------------
% Custom environments (shared with the core specification)
% ---------------------------------------------------------------------------
\newtcolorbox{openquestion}[1][]{
  enhanced, breakable,
  colback=epiphanymist, colframe=epiphanycrimson,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Open Question},
  coltitle=white, colbacktitle=epiphanycrimson,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}
\newtcolorbox{rationale}[1][]{
  enhanced, breakable,
  colback=epiphanymist, colframe=epiphanyteal,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Rationale},
  coltitle=white, colbacktitle=epiphanyteal,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}
% Numbered within chapter (this document has chapters); see core_spec.tex's
% requirement box for why a plain counter + `code=` step is used instead of
% tcolorbox's own "auto counter, number within=..." keys.
\newcounter{requirement}[chapter]
\renewcommand{\therequirement}{\thechapter.\arabic{requirement}}
\newtcolorbox{requirement}[1][]{
  enhanced, breakable,
  colback=white, colframe=epiphanygold,
  fonttitle=\bfseries\color{white}, code={\refstepcounter{requirement}}, title={\scshape\hspace{2pt}Requirement~\therequirement},
  coltitle=white, colbacktitle=epiphanygold,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}
\newtcolorbox{nongoal}[1][]{
  enhanced, breakable,
  colback=epiphanymist, colframe=epiphanyslate,
  fonttitle=\bfseries\color{white}, title={\scshape\hspace{2pt}Non-Goal},
  coltitle=white, colbacktitle=epiphanyslate,
  arc=1pt, boxrule=0pt, leftrule=2pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  attach boxed title to top left={xshift=0pt, yshift=0pt},
  boxed title style={arc=0pt, sharp corners, boxrule=0pt, left=6pt, right=8pt, top=2pt, bottom=2pt},
  #1
}

\newcommand{\MUST}{\textbf{MUST}}
\newcommand{\MUSTNOT}{\textbf{MUST}\nobreak\ \textbf{NOT}}
\newcommand{\SHOULD}{\textbf{SHOULD}}
\newcommand{\SHOULDNOT}{\textbf{SHOULD}\nobreak\ \textbf{NOT}}
\newcommand{\MAY}{\textbf{MAY}}

\setlist[itemize]{topsep=2pt, itemsep=3pt, parsep=0pt}
\setlist[enumerate]{topsep=2pt, itemsep=3pt, parsep=0pt}
\setlist[description]{topsep=2pt, itemsep=5pt, parsep=0pt}
\AtBeginDocument{\color{epiphanyink}}

% ---------------------------------------------------------------------------
% Document
% ---------------------------------------------------------------------------
\begin{document}

\begin{titlepage}
  \thispagestyle{empty}
  \centering
  \vspace*{2.2in}
  {\color{epiphanygold}\rule{3in}{0.8pt}}\\[18pt]
  {\titlefont\fontsize{34pt}{38pt}\selectfont\color{epiphanyteal}\bfseries Epiphany}\\[10pt]
  {\Large\scshape\color{epiphanyslate}Text Projection}\\[6pt]
  {\large\itshape\color{epiphanyslate}A companion to the Core Specification}\\[14pt]
  {\color{epiphanygold}\rule{3in}{0.8pt}}\\[24pt]
  {\normalsize\color{epiphanyink}Version 0.7.0 --- Canonical blob rejection and single-version header gating}\\[4pt]
  {\small\color{epiphanyslate}Normative for the text form it defines}
  \vfill
\end{titlepage}

\tableofcontents

% ===========================================================================
\chapter{About This Companion}
\label{ch:about}

The \emph{Text Projection} is a companion to the Epiphany Core Specification. It
fulfils the delegation the core specification makes in Chapter~8,
\sectionsc{Text Projection} (\texttt{sec:format:textproj}), which declares that
``the format admits a deterministic projection to a canonical s-expression text
form'' and that ``the text projection is normative'' --- while leaving the form
itself unwritten. The Binary Format companion likewise excludes it: ``the
canonical s-expression form --- that is the \emph{Text Projection} companion's''.

This document supplies that form.

\section{What This Document Covers}

\begin{itemize}
  \item The canonical text syntax: atoms, byte strings, text, and the layout that
    makes the projection deterministic.
  \item What is projected, and what is deliberately not.
  \item The projection and parse requirements, including the bidirectional
    round-trip the core specification demands.
\end{itemize}

It does \emph{not} cover the binary encoding of anything --- that is the Binary
Format companion's --- nor the semantics of any operation, which is the Operation
Catalog's. The projection is a \emph{re-presentation} of the canonical document,
never a second definition of it. Wherever the two could disagree, the binary form
is normative and the projection is wrong.

\section{The Subject of the Projection}

The projection's subject is the \textbf{canonical document}, not the file. Per the
core specification's Chapter~8 \sectionsc{Text Projection}, the projection
\MUSTNOT{} be required to preserve chunk offsets, compression choices, cache
chunks (operation indexes, layout caches, integrity indexes), garbage bytes from
prior commits, or superblock generation numbers, slot assignments, and CRCs.

Consequently two bundles that differ only in physical layout project to the same
text, and a text re-serializes to \emph{a} bundle rather than to \emph{the}
bundle it came from. That is the intent, not a limitation: the physical file is
an encoding of the document, and the projection is of the document.

A bundle \MAY{} cache its own projection in a \texttt{TextProjection} chunk named
by \texttt{Manifest.text\_projection\_root}. That chunk is a \textbf{non-canonical
accelerator} (core specification Chapter~8, \sectionsc{Schema Versioning}): a
reader need not understand it, and a writer preserves it verbatim or discards it.
A cached projection that disagrees with the operations it claims to project is
\emph{stale}, not authoritative.

% ===========================================================================
\chapter{The Canonical Text Form}
\label{ch:form}

\section{Encoding and Character Set}

\begin{requirement}
  \label{req:textproj:charset}
  A text projection \MUST{} be UTF-8, with no byte-order mark. Every text field
  it carries \MUST{} be in Unicode NFC, matching the canonical-text rule of the
  core specification's Appendix~D, \sectionsc{Text and Unicode}. A parser
  \MUST{} reject non-NFC text rather than normalize it.
\end{requirement}

\section{Atoms}

An \emph{atom} is a symbol, an integer, a byte string, or a text string.

\textbf{Symbols} are lowercase ASCII words, possibly hyphenated:
\texttt{envelope}, \texttt{insert-event}, \texttt{strict-inverse}. They name
constructors and enumeration cases. A symbol is never quoted.

\textbf{Integers} are written in base ten, with a leading \texttt{-} for negative
values, no leading zeros, and no leading \texttt{+}. Zero is \texttt{0}, never
\texttt{-0}.

\textbf{Byte strings} carry every identifier, hash, and opaque payload.

\begin{requirement}
  \label{req:textproj:hex}
  A byte string \MUST{} be written as \texttt{\#x} followed by an even number of
  \textbf{lowercase} hexadecimal digits, one pair per byte, in the order the
  bytes appear in the canonical binary form. The empty byte string is
  \texttt{\#x}. A parser \MUST{} reject uppercase digits, an odd digit count, and
  any separator within the digits.
\end{requirement}

\begin{rationale}
  One rule for every byte string. Hexadecimal has no alphabet variant and no
  padding to canonicalize, it is greppable, and a corrupted character is locally
  obvious. Identifiers are 16 or 32 bytes, so its expansion costs nothing where
  it is read; only an inlined snapshot pays. The core specification permits
  ``base64 \emph{or another canonical text form}''; a second encoding would buy
  a quarter of the bytes of the one body nobody reads, at the price of pinning an
  alphabet, a padding rule, and a line-wrapping rule, and of choosing which
  encoding applies where. Ratified at 0.1.0.
\end{rationale}

\textbf{Text strings} are double-quoted. Inside a string, \texttt{\textbackslash{}"}
denotes a quotation mark, \texttt{\textbackslash{}\textbackslash{}} a backslash,
\texttt{\textbackslash{}n} a line feed, and \texttt{\textbackslash{}t} a tab; no
other escape exists.

\begin{requirement}
  \label{req:textproj:string-escapes}
  A text string \MUST{} escape exactly the characters that require it: the
  quotation mark, the backslash, U+000A, and U+0009. Every other character
  \MUST{} appear literally. A parser \MUST{} reject an escape sequence outside
  this set, and \MUST{} reject a literal character that the writer was required
  to escape.
\end{requirement}

\begin{rationale}
  ``Escape exactly'' rather than ``escape at least'': the text is canonical, so
  two spellings of one string cannot both be valid. This is the same injectivity
  the binary form rests on (Binary Format,
  \texttt{req:binfmt:decode-vectors}), stated for text.
\end{rationale}

\section{Layout}

\begin{requirement}
  \label{req:textproj:envelope-per-line}
  A projection is a sequence of lines separated by a single U+000A, with a final
  U+000A and no other trailing whitespace. Each line is one complete
  s-expression. Tokens within a line are separated by exactly one space; there is
  no other whitespace, and no indentation.

  Each operation envelope \MUST{} occupy exactly one line.
\end{requirement}

\begin{rationale}
  The core specification's stated use case is that ``merge conflicts surface at
  the operation-envelope level, which is the meaningful level for collaborative
  editing''. One envelope per line makes a line-based three-way merge conflict
  \emph{exactly} an envelope conflict --- never a conflict inside an envelope,
  which could otherwise produce a syntactically valid operation that neither side
  wrote. It also disposes of indentation: there is no whitespace to canonicalize,
  so the ``identical semantics project to identical text'' requirement below has
  nothing to hide in.

  The lines are long. Readability is a \emph{tooling} concern, and a
  pretty-printer is free to reformat for display; what it must not do is write
  the reformatted text back and call it a projection. Ratified at 0.1.0.
\end{rationale}

% ===========================================================================
\chapter{What Is Projected}
\label{ch:content}

\section{Derive, or Carry --- Never Both}
\label{sec:content:derive-or-carry}

The manifest's references are \emph{physical}. A \texttt{ChunkRef} carries an
offset, a compressed length, and a compression algorithm; a \texttt{BlobRef}
carries the same. Those are exactly the things the projection \MUSTNOT{}
preserve. But a reference also carries an \emph{identity} --- a chunk id, a
content hash, a blob id --- and every one of those is a function of the content.

One rule resolves both, and it is the rule
Requirement~\ref{req:textproj:reduced-state-derived} already applies to reduced
state.

\begin{requirement}
  \label{req:textproj:derive-or-carry}
  A projection \MUST{} carry exactly what the document does not determine, and
  \MUSTNOT{} carry anything it does.

  \begin{itemize}
    \item \textbf{Physical attributes} --- a chunk's or blob's \texttt{offset},
      \texttt{compressed\_length}, and \texttt{compression}, and a chunk's
      \texttt{uncompressed\_length} --- \MUSTNOT{} appear. A serializer chooses
      them freely.
    \item \textbf{Derivable identities} --- \texttt{ChunkId},
      \texttt{ContentHash}, \texttt{BlobId} --- \MUSTNOT{} appear. They are
      re-derived from the content by the derivations the Binary Format companion
      pins (\sectionsc{Content Hashing}, \sectionsc{Domain-Separated Preimages}).
    \item \textbf{Non-derivable identities} \MUST{} appear. In schema major~0
      there is exactly one: \texttt{SnapshotId}, which the Binary Format
      companion declares opaque, with readers forbidden from deriving it
      (\texttt{req:binfmt:snapshot-id-opaque}).
    \item \textbf{Content and semantic attributes} \MUST{} appear: a chunk's
      \texttt{kind} and \texttt{schema\_version} and its uncompressed payload; a
      blob's media type, declared maximum uncompressed length if any, and its
      payload.
  \end{itemize}

  A parser \MUST{} reject a projection carrying a value this requirement forbids.
\end{requirement}

\begin{rationale}
  Carrying a derivable identity would reproduce, at the level of a chunk, the
  defect Requirement~\ref{req:textproj:reduced-state-derived} rules out at the
  level of the document: two sources of truth for one fact, with nothing to stop
  them disagreeing. Carrying a physical attribute would make two encodings of one
  document project to two texts, breaking
  Requirement~\ref{req:textproj:canonical-text}.

  \texttt{SnapshotId} is the sole exception, and it is an exception for a stated
  reason rather than an oversight: v0 has no snapshot producer, so the identity
  has nothing to derive \emph{from}, and the Binary Format companion accordingly
  pins it as sixteen opaque bytes that a reader \MUSTNOT{} attempt to verify. A
  projection must carry what it cannot recompute.
\end{rationale}

\section{Document Structure}
\label{sec:content:structure}

A projection is, in order:

\begin{enumerate}
  \item a \texttt{(text-projection <version>)} header line, naming the version of
    \emph{this companion} the text conforms to;
  \item a \texttt{(document \#x<document-id>)} line, and a
    \texttt{(lineage \#x<lineage-id>)} line if the manifest declares one;
  \item zero or more \texttt{(profile ...)} lines, in canonical order;
  \item zero or more \texttt{(extension ...)} lines, in canonical order;
  \item at most one \texttt{(canonical-base ...)} line;
  \item zero or more \texttt{(blob ...)} lines, in canonical order;
  \item zero or more \texttt{(envelope ...)} lines, in canonical operation order
    (core specification Appendix~D).
\end{enumerate}

\begin{requirement}
  \label{req:textproj:header-version}
  A parser implementing this companion \MUST{} accept exactly one header
  version: \texttt{(0 7 0)}, the version of the companion it implements. It
  \MUST{} reject any other version at line one.

  Multi-version acceptance and text migrate-on-read are deferred in the same
  posture as op-payload migrate-on-read: support belongs in an explicit,
  version-keyed migration path when a real consumer requires it. The current
  parser \MUSTNOT{} speculate by accepting another version.
\end{requirement}

Every sequence is written in the normative order its binary counterpart uses. The
projection introduces no ordering of its own.

\section{Canonical Blobs}
\label{sec:content:blobs}

\begin{requirement}
  \label{req:textproj:canonical-blobs}
  A blob referenced by a canonical operation or by canonical reduced state is
  itself canonical (core specification Chapter~8,
  \sectionsc{Canonical and Non-Canonical Roots}). Every such blob \MUST{} be
  projected as a \texttt{(blob ...)} line carrying its media type, its declared
  maximum uncompressed length if it declares one, and its uncompressed payload.

  Its \texttt{BlobId}, content hash, offset, lengths, and compression are
  re-derived (Requirement~\ref{req:textproj:derive-or-carry}). The blob lines are
  ordered and de-duplicated by their projected form
  (Requirement~\ref{req:textproj:derived-ordering}), not by the binary order,
  which reads the offset.

  A blob referenced only by acceleration structures is non-canonical and
  \MUSTNOT{} be projected.

  Parser handling of unreferenced blob lines is specified by
  Requirement~\ref{req:textproj:reject-unreferenced-blobs}.
\end{requirement}

\begin{rationale}
  An embedded image, font, or audio recording that a canonical operation
  references is part of the document. Omitting it would make the projection lossy
  for exactly the documents most in need of archival --- and lossy \emph{silently},
  since the operations that reference the blob would still be there, pointing at a
  blob id the text no longer contains. This requirement was absent from version
  0.1.0 of this companion, and from the core specification's own list of what the
  projection preserves; both are corrected.
\end{rationale}

\begin{requirement}
  \label{req:textproj:reject-unreferenced-blobs}
  A parser \MUST{} reject every \texttt{(blob ...)} line whose blob is
  unreferenced by canonical state
  (Requirement~\ref{req:textproj:canonical-blobs}). At companion
  version~0.7.0, neither a canonical operation nor canonical reduced state can
  carry a \texttt{BlobId}; canonical state therefore cannot reference a blob,
  and a parser \MUST{} reject every \texttt{(blob ...)} line.
\end{requirement}

\begin{rationale}
  A blob-bearing text at this version is necessarily non-canonical. Accepting
  one stages a blob into the bundle that the next projection silently drops,
  causing data loss and falsifying
  $\textrm{project}(\textrm{serialize}(\textrm{parse}(T))) = T$ for that text.
  Forward compatibility belongs to header-version gating
  (Requirement~\ref{req:textproj:header-version}), not to leniency here.
\end{rationale}

\section{Profile Declarations}
\label{sec:content:profiles}

A \texttt{(profile ...)} line carries the profile's identity, its semantic
version, and its constraints.

\begin{requirement}
  \label{req:textproj:profile-id}
  A profile identity is a symbol for each closed-vocabulary profile
  (\texttt{full}, \texttt{read-only}, \texttt{lite}), and
  \texttt{(custom \#x<registry-id>)} for \texttt{ProfileId::Custom}, whose
  registry id is sixteen bytes.
\end{requirement}

\begin{rationale}
  Version 0.1.0 required a symbol for every profile, which made a custom profile
  unrepresentable and its claim to preserve ``all profile declarations'' false.
\end{rationale}

\section{Extension Declarations}
\label{sec:content:extensions}

\begin{requirement}
  \label{req:textproj:extension-declaration}
  An \texttt{(extension ...)} line \MUST{} carry every field of the declaration:
  its identity, its semantic version, whether it is required, its affected object
  kinds, its edit barriers, and its preserved chunk roots. The affected object
  kinds and the edit barriers are opaque to the bundle and are projected as byte
  strings, verbatim.

  Fields appear in the ratified declaration order (core specification Chapter~8,
  \sectionsc{Extension Declarations}).

  Each preserved chunk root is projected as its \texttt{kind}, its
  \texttt{schema\_version}, and its uncompressed payload
  (Requirement~\ref{req:textproj:derive-or-carry}), never as a
  \texttt{ChunkRef}: a \texttt{ChunkRef} is a physical reference, and the
  projection has no file to point into. The chunk list is ordered and
  de-duplicated by projected form
  (Requirement~\ref{req:textproj:derived-ordering}), because
  \texttt{ChunkRef}'s binary order breaks ties on the offset.
\end{requirement}

\begin{rationale}
  \texttt{affected\_object\_kinds} and \texttt{edit\_barriers} have ratified
  structured shapes (\texttt{ObjectKind}, \texttt{EditBarrier}) \emph{and}
  canonical byte encodings, and the bundle stores them opaquely: it preserves
  them across reads and writes without interpreting them. At 0.3.0 the projection
  does the same, carrying their canonical bytes, on the principle that the
  projection interprets nothing the bundle does not.

  This is a deferral, not a conclusion. A later revision \MAY{} project them
  structurally under
  Requirement~\ref{req:textproj:value-projection}; because their canonical bytes
  are unchanged by that, doing so does not change the document, only its text ---
  and that is a version-gated change to this companion, not to the format.
\end{rationale}

\begin{rationale}
  Version 0.1.0 carried only identity, the required flag, barriers, and a list of
  chunks, dropping the semantic version and the affected object kinds outright and
  leaving the chunks' representation undefined. An extension's declaration governs
  how the canonical document is \emph{interpreted}; a projection that loses part
  of it does not determine the document.
\end{rationale}

\section{Ordering What the Binary Form Ordered Physically}
\label{sec:content:derived-ordering}

``Keep the binary form's order'' is available only where the binary order is a
function of data the projection preserves. Two sequences fail that test.

\begin{requirement}
  \label{req:textproj:derived-ordering}
  Where a sequence's binary order depends on attributes
  Requirement~\ref{req:textproj:derive-or-carry} erases, the projection \MUST{}
  order its elements by their \textbf{projected form}, ascending, comparing the
  UTF-8 bytes of the rendered element; and \MUST{} emit at most one element per
  distinct projected form.

  In schema major~0 this applies to exactly two sequences:

  \begin{itemize}
    \item the \texttt{(blob ...)} lines, whose binary counterpart
      \texttt{blob\_roots} is sorted by the full \texttt{BlobRef} encoding ---
      which contains the offset, the compressed length, and the compression
      algorithm; and
    \item an extension's preserved chunk roots, sorted in binary by
      \texttt{ChunkRef}'s order, whose key is the kind, then the content hash,
      then the \textbf{offset}, with the compressed length, the uncompressed
      length, and the compression as further tie-breakers.
  \end{itemize}

  Every other projected sequence keeps the binary order, because every other
  binary order reads only preserved data. In particular the profile and extension
  declarations are sorted in binary by the semantic keys
  $(\texttt{profile\_id}, \texttt{version})$ and
  $(\texttt{extension\_id}, \texttt{version})$, and the envelopes by canonical
  operation order.
\end{requirement}

\begin{rationale}
  Two bundles that differ only in physical layout are the \emph{same document},
  and Requirement~\ref{req:textproj:canonical-text} obliges them to project to
  byte-identical text. Inheriting the binary order for these two sequences would
  let a chunk's file offset decide the order of the text --- so relocating a chunk,
  which changes no semantics, would change the projection. That is precisely the
  failure the requirement forbids.

  The de-duplication is the same point from the other side. A chunk is
  content-addressed: two entries with identical kind, schema version, and payload
  \emph{are} one chunk, and appear twice only because the writer stored the bytes
  twice. A blob with identical media type, declared maximum, and payload is one
  blob. Their projected forms are identical, so the binary form's distinction
  between them is a physical fact --- and a duplicated line would smuggle that
  physical fact into a text that claims to have erased it. Emitting one line is
  not a loss; it is the erasure working.

  Ordering by the projected form is total and deterministic, and it reads nothing
  but what is projected. For chunks it coincides with ordering by the derived
  \texttt{ChunkId}, since that id is a function of exactly the kind, schema, and
  payload the line carries --- which is a pleasing check that the rule is reading
  the right thing.
\end{rationale}

\section{Projecting Canonical Values}
\label{sec:content:values}

An operation payload embeds canonical values from the core specification's
Chapter~5 --- an \texttt{Event}, a \texttt{Pitch}, a \texttt{Region}, a
\texttt{TimeSignature}. This document does \emph{not} restate their shapes. It
states one rule for turning any of them into text, and the shapes stay where they
are ratified.

\begin{requirement}
  \label{req:textproj:operation-vocabulary}
  The core specification's Chapter~6 operation vocabulary --- the envelope, its
  stamp and causal context, the payload, the operation kinds, and their
  sub-vocabularies --- \MUST{} be projected by the productions of
  Chapter~\ref{ch:grammar}, not by
  Requirement~\ref{req:textproj:value-projection}. The value-projection
  rule governs exactly the \texttt{value} positions those productions contain,
  and no other positions.

  An operation kind's payload record \MUST{} be inlined into its production. The
  record exists so each variant can name a type; it is not a modelling
  distinction, and the binary form agrees: \texttt{OperationKind}'s encoding
  writes the kind tag and then delegates to the record, adding no bytes for the
  wrapper. It adds no text here either, for the same reason clause~2 of
  Requirement~\ref{req:textproj:value-projection} makes a newtype transparent.
\end{requirement}

\begin{requirement}
  \label{req:textproj:schema-directed}
  The projection is \textbf{schema-directed}. At every position a reader knows
  the type it expects, from the ratified schema and from the position itself,
  exactly as the binary decoder does. The grammar of
  Chapter~\ref{ch:grammar} describes the \emph{shape} of the text; it is not a
  standalone unambiguous language, and a reader \MUSTNOT{} attempt to recover a
  value's type from its shape.

  Three consequences follow, and a reader resolves each by the type it expects:

  \begin{itemize}
    \item \texttt{()} is the empty sequence, and it is the absent option.
    \item A bare symbol is a fieldless variant, and it is a struct with no
      fields.
    \item A parenthesised list whose first element is a symbol is a struct, and
      it is a sequence whose first element is a symbol.
  \end{itemize}
\end{requirement}

\begin{rationale}
  This states what the ratified rules already require rather than adding a new
  constraint. A struct is \texttt{(<type-name> <field>\ldots)} and a sequence is
  a parenthesised list of its elements; a sequence whose first element is a
  fieldless variant is therefore shape-identical to a struct. The collision is
  \emph{irreducible} without new syntax, and it is reachable from the first
  pitched note a score contains: a \texttt{PitchedEvent} carries
  \texttt{articulations} and \texttt{ornaments}, sequences of a zero-field
  \texttt{ArticulationMark} and \texttt{OrnamentMark}, alongside an optional
  \texttt{DynamicMark} --- so one \texttt{insert-event} line holds an empty
  sequence and an absent option, both spelled \texttt{()}, and a two-element
  sequence spelled like a two-field struct.

  Making the text standalone-unambiguous would buy nothing, because
  Requirement~\ref{req:textproj:strict-parse} already obliges a parser to reject
  ``a duplicate in a set-typed field'' --- which it cannot do without knowing that
  the field is set-typed. The parser consults the schema either way. Spending
  syntax to remove a shape ambiguity, while leaving the schema dependency it was
  meant to remove, would make every line longer and no tool simpler.

  The binary form is schema-directed for the same reason and pays the same price:
  its bytes do not say what they are either.
\end{rationale}

\begin{requirement}
  \label{req:textproj:value-projection}
  A canonical value is projected thus:

  \begin{enumerate}
    \item A \textbf{struct} becomes \texttt{(<type-name> <field>\ldots)}, where
      the type name is the core specification's name in lower-case
      hyphenated form and the fields appear \emph{positionally}, in the order
      that specification's ratified listing declares them. Field names are not
      written. A struct with \emph{no} fields is the bare symbol
      \texttt{<type-name>}, as a fieldless variant is; it encodes to no bytes in
      the binary form and carries no value here.
    \item A \textbf{newtype} --- a struct of exactly one unnamed field ---
      is projected as that field alone, with no wrapper. This mirrors the binary
      form, in which a newtype delegates to its field and adds no bytes.
    \item A \textbf{tagged union} becomes
      \texttt{(<variant-name> <field>\ldots)}. A variant with no fields is the
      bare symbol \texttt{<variant-name>}.
    \item An \textbf{option} is \texttt{()} when absent and
      \texttt{(some <value>)} when present.
    \item A \textbf{sequence}, \textbf{set}, or \textbf{map} is a parenthesised
      list of its elements, \emph{in the order the binary form writes them},
      except where Requirement~\ref{req:textproj:derived-ordering} applies; a
      map entry is \texttt{(<key> <value>)}. An empty one is \texttt{()}, which
      Requirement~\ref{req:textproj:schema-directed} distinguishes from an absent
      option by the expected type. The projection invents no ordering of its own,
      and a set that the binary form writes strictly increasing is written
      strictly increasing here
      (Requirement~\ref{req:textproj:strict-parse}).

      A \textbf{byte string} is not a sequence. Where the binary form writes a
      length-prefixed run of bytes rather than a counted sequence of elements ---
      an opaque extension payload, a \texttt{SoundConfiguration} --- the
      projection writes a byte string, not a list of integers.
    \item \textbf{Leaves.} An identifier or hash is a byte string. An integer is
      an integer. A boolean is \texttt{true} or \texttt{false}. Canonical text is
      a quoted string. A rational is \texttt{(ratio <numerator> <denominator>)},
      in lowest terms with a positive denominator and the sign on the numerator;
      zero is \texttt{(ratio 0 1)}. A \texttt{CanonicalF64} is a byte string of
      its eight canonical little-endian IEEE~754 bytes.
  \end{enumerate}
\end{requirement}

\begin{rationale}
  \textbf{One rule, not forty productions.} Spelling out a production per value
  type would restate the entire Chapter~5 data model in a second normative
  document, and two normative listings of one struct is precisely the drift this
  project has already been bitten by. A rule cannot drift from the listing it
  reads.

  \textbf{Positional fields.} The declaration order is already normative --- the
  binary form depends on it --- so field names would be redundant, would double
  the length of every line, and would create a second thing to keep in step with
  a rename. The constructor name carries the context a reader needs.

  \textbf{Newtypes are transparent} for the same reason they are transparent in
  the binary form: a \texttt{MusicalPosition} \emph{is} a rational, and wrapping
  it would put a distinction in the text that the document does not make.

  \textbf{Floats are bytes, never decimal.} A decimal rendering of an
  \texttt{f64} is not canonically unique --- shortest-round-trip and
  seventeen-significant-digit forms both round-trip, and $-0.0$ has two
  spellings --- so a decimal float would break
  Requirement~\ref{req:textproj:canonical-text} at the first tempo mark. The core
  specification already forbids computed floats in canonical state and stores the
  eight bytes; the projection carries those eight bytes. This costs readability
  in exactly one place, and buys canonicality everywhere.
\end{rationale}

\section{Reduced State}

\begin{requirement}
  \label{req:textproj:reduced-state-derived}
  The projection \MUST{} preserve canonical reduced state \emph{by preserving the
  operations that determine it}. It \MUSTNOT{} carry a second, literal copy of
  the reduced state.
\end{requirement}

\begin{rationale}
  Reduced state is a deterministic function of the operation set and the
  canonical base (core specification Chapter~6, \sectionsc{Design Principles}).
  A text that carried both would have two sources of truth for one fact, and
  nothing could stop them disagreeing --- a projection with an internally
  contradictory document is worse than no projection. This reading is what the
  core specification's own round-trip clause already implies: text that ``parses
  to identical canonical document semantics'' must re-serialize to bundles with
  identical semantics, and reduced state is a function of semantics.

  Read the core specification's ``all canonical reduced state'' as
  \emph{determines}, not \emph{contains}. Ratified at 0.1.0.
\end{rationale}

\section{The Canonical Base Snapshot}

\begin{requirement}
  \label{req:textproj:base-snapshot-inline}
  If the manifest declares a canonical base, the projection \MUST{} carry a
  \texttt{(canonical-base ...)} line bearing:

  \begin{itemize}
    \item the \texttt{SnapshotId}, verbatim --- the one identity that is carried
      rather than derived (Requirement~\ref{req:textproj:derive-or-carry});
    \item the causal frontier it materializes, as opaque bytes;
    \item its reduction-algorithm version;
    \item the profile under which it was produced;
    \item its root chunk's \texttt{schema\_version}; and
    \item \emph{the root chunk's uncompressed payload, inline}, as a single byte
      string. That payload is the canonical byte form of the reduced state the
      snapshot materializes.
  \end{itemize}

  The root chunk's kind is \texttt{Snapshot} by role and is not written. The
  \texttt{SnapshotRef}'s \texttt{root} and \texttt{hash} are \textbf{re-derived}:
  the chunk's content hash is
  $\textrm{hash}(\texttt{Snapshot}, \textit{schema}, \textit{payload})$ under the
  Binary Format companion's chunk-hash preimage, and it is both the root's
  \texttt{ChunkId} and the \texttt{SnapshotRef}'s \texttt{hash}. A parser
  \MUST{} perform that derivation rather than read it from the text.
\end{requirement}

\begin{rationale}
  A canonical base exists precisely so that the operations before its frontier
  need not be retained. Where they have been pruned, the snapshot is \emph{not}
  derivable from anything else in the document, and a projection that carried
  only a reference would be \textbf{lossy} --- the text would no longer determine
  the document, which is the one thing it is for. The core specification permits
  the payload to be ``encoded compactly or referenced externally''; inline and
  compact is the choice that keeps the text self-contained.

  The snapshot diffs as one opaque atom. That is acceptable: merges happen among
  operations, and a base snapshot changes only when the document is compacted, at
  which point the whole line changes anyway. A later revision \MAY{} project the
  snapshot structurally; doing so does not break the round trip, because the
  document it denotes is unchanged. Ratified at 0.1.0.

  Schema major~0 has no snapshot producer --- pruning and canonical-base creation
  are deferred (Binary Format, \sectionsc{SnapshotId}) --- so this requirement
  binds whoever writes the first one, and cannot be exercised before then.
\end{rationale}

% ===========================================================================
\chapter{Requirements}
\label{ch:requirements}

\section{Canonicality}

\begin{requirement}
  \label{req:textproj:canonical-text}
  Two bundles whose canonical document semantics are identical \MUST{} project to
  \textbf{byte-identical} text. A projector \MUSTNOT{} have any freedom the
  document does not determine: no optional whitespace, no alternative spelling of
  an atom, no ordering choice.
\end{requirement}

\section{Round Trip}

\begin{requirement}
  \label{req:textproj:roundtrip}
  Parsing a projection and re-serializing it to binary \MUST{} yield a bundle
  whose canonical document semantics are identical to the original's. The
  bundle's physical layout, chunking, and compression \MAY{} differ.

  Equivalently, and more usefully to an implementer: for every bundle $B$,
  \[
    \textrm{semantics}(\textrm{parse}(\textrm{project}(B))) =
    \textrm{semantics}(B),
  \]
  and for every valid projection $T$,
  \[
    \textrm{project}(\textrm{serialize}(\textrm{parse}(T))) = T .
  \]
  The second equation is the text's own injectivity: it is \emph{stronger} than
  the first, and it is the one a conformance test can check with byte equality.
\end{requirement}

\section{Strict Parsing}

\begin{requirement}
  \label{req:textproj:strict-parse}
  A parser \MUST{} reject any text that is not the canonical projection of the
  document it denotes. It \MUSTNOT{} normalize: not whitespace, not letter case
  in a byte string, not an escape sequence, not an out-of-order sequence, not a
  duplicate in a set-typed field.

  Accepting non-canonical text and normalizing it \emph{is} accepting it, and
  does not satisfy this requirement.

  Rejecting a duplicate in a set-typed field requires knowing that the field is
  set-typed. A parser therefore reads the schema
  (Requirement~\ref{req:textproj:schema-directed}); shape alone never suffices.
\end{requirement}

\begin{rationale}
  This is the same discipline the binary decoders carry, and it exists for the
  same reason: a lenient parser makes two texts denote one document, and the
  projection's contract is that a text \emph{determines} its document. The Binary
  Format companion learned this concretely --- a whole-value re-encode guard
  catches the fields a decoder normalizes and is blind to order-preserving
  sequences, and a guard on an outer value can \emph{mask} a lenient inner codec
  rather than fix it (Binary Format,
  \sectionsc{The Decode Vector Corpus}). A text parser inherits both hazards, and
  the cheapest total defence is the same one: re-project the parsed document and
  compare, \emph{and} check per-site the orders that re-projection would restore.
\end{rationale}

\section{Conformance}

\begin{requirement}
  \label{req:textproj:conformance}
  An implementation claiming Text Projection conformance \MUST{} implement both
  directions. A projector alone does not conform: the round trip
  (Requirement~\ref{req:textproj:roundtrip}) is the requirement, and half of it
  is not checkable.
\end{requirement}

% ===========================================================================
\chapter{Grammar}
\label{ch:grammar}

\begin{lstlisting}
; Notation. X* is zero or more X, X? is zero or one. Within a line, adjacent
; elements of a repetition are separated by exactly one U+0020, and an empty
; repetition contributes nothing -- so "(" value* ")" spells () when empty. The
; line productions below carry their own trailing LF, and are simply
; concatenated. LF is U+000A. Terminals are quoted; U+XXXX names a codepoint.

projection     ::= header document lineage? profile* extension*
                 canonical-base? blob* envelope*

header         ::= "(text-projection " version ")" LF
version        ::= "(" integer " " integer " " integer ")"

document       ::= "(document " bytes ")" LF
lineage        ::= "(lineage " bytes ")" LF

profile        ::= "(profile " profile-id " " version " " constraints ")" LF
profile-id     ::= "full" | "read-only" | "lite" | "(custom " bytes ")"
constraints    ::= "(constraints " integer " " retention ")"
retention      ::= "(retention " integer " " option " " bool ")"

extension      ::= "(extension " bytes " " version " " bool
                 " (" chunk* ") " bytes " " bytes ")" LF
                 ; id, version, required, chunks, affected-kinds, barriers
                 ; (the ratified declaration order)
chunk          ::= "(chunk " chunk-kind " " schema " " bytes ")"
chunk-kind     ::= "operation-envelope-block" | "operation-index" | "snapshot"
               | "blob" | "extension-data" | "text-projection"
               | "layout-cache" | "integrity-index" | "manifest"
schema         ::= "(schema " integer " " integer ")"

canonical-base ::= "(canonical-base " bytes " " bytes " " integer
                 " " profile-id " " schema " " bytes ")" LF
                 ; snapshot-id, frontier, reduction version, profile,
                 ; root schema, root payload

blob           ::= "(blob " string " " option " " bytes ")" LF
                 ; media type, declared max uncompressed length, payload

envelope       ::= "(envelope " bytes " " bytes " " stamp " " causal
                 " " option " " payload ")" LF
                 ; id, author, stamp, causal context, transaction, payload
stamp          ::= "(stamp " integer " " integer " " bytes ")"
causal         ::= "(causal (" replica-seen* ") (" bytes* "))"
replica-seen   ::= "(" bytes " " integer ")"

payload        ::= "(primitive " kind ")"
               | "(resolve-conflict " bytes " " action ")"
               | "(undo " bytes " " policy ")"
               | "(resolve-equivocation " bytes " " bytes ")"

action         ::= "accept-loser" | "keep-winner" | "dismiss"
               | "(override " bytes ")" | "(reanchor " bytes ")"
               | "(registered " bytes ")"
policy         ::= "strict-inverse" | "best-effort" | "cascade"

; --- Operation kinds inline their payload records as required by
; --- req:textproj:operation-vocabulary. Fields are the Operation Catalog's
; --- payload schema, positionally, in declaration order. Embedded Chapter-5
; --- values occur only at <value> positions and follow
; --- req:textproj:value-projection.

kind           ::= "(insert-event " bytes " " value ")"
               | "(delete-event " bytes " " tuplet-comp ")"
               | "(respell-pitch " bytes " " value ")"
               | "(create-cross-cutting " cross-cutting ")"
               | "(change-region-time-model " bytes " " value
                     " (" bytes* ") " remapping ")"
               | "(set-user-system-break " bytes " " value " " bool ")"
               | "(declare-transaction " bytes " " string " " option ")"
               | "(registered " bytes " " bytes ")"
               | "(modify-event " value ")"
               | "(transpose (" bytes* ") " integer ")"
               | "(insert-identified-pitch " bytes " " value ")"
               | "(delete-identified-pitch " bytes ")"
               | "(modify-identified-pitch " bytes " " value ")"
               | "(delete-cross-cutting " bytes ")"
               | "(modify-cross-cutting " cross-cutting ")"
               | "(create-region " value ")"
               | "(delete-region " bytes ")"
               | "(create-staff-instance " bytes " " value ")"
               | "(delete-staff-instance " bytes ")"
               | "(create-voice " bytes " " value ")"
               | "(delete-voice " bytes ")"
               | "(set-metadata " value ")"
               | "(set-metric-grid " bytes " " option ")"
               | "(set-user-page-break " bytes " " value " " bool ")"
               | "(create-staff " value ")"
               | "(set-time-signature " bytes " " value " " option ")"
               | "(set-tempo-segment " option " " value " " option ")"
               | "(set-staff-layout " bytes " " option " " option " " bool ")"
               | "(create-repeat-structure " value ")"
               | "(delete-repeat-structure " bytes ")"
               | "(transpose-interval (" bytes* ") " value ")"

tuplet-comp    ::= "not-in-tuplet" | "(replace-with-rest " value ")"
               | "(rewrite-tuplets (" bytes* "))"
               | "(cascade-delete-tuplets (" bytes* "))"
cross-cutting  ::= "(tie " value ")" | "(slur " value ")"
               | "(beam " value ")" | "(spanner " value ")"
remapping      ::= "preserve-time" | "(reassign (" reassign-entry* "))"
reassign-entry ::= "(" bytes " " ratio ")"          ; event id, musical position

; --- Values and leaves.

; A struct, a sequence, a set, a map and an option all have this one shape. The
; reader tells them apart by the type it expects, never by the shape:
; req:textproj:schema-directed. Meaning is req:textproj:value-projection.
value          ::= "(" value* ")"
               | symbol | bytes | integer | bool | string | ratio | option
option         ::= "()" | "(some " value ")"
ratio          ::= "(ratio " integer " " integer ")"

bytes          ::= "#x" hexdigit*        ; even count, lowercase
integer        ::= "-"? digit+           ; no leading zeros, no "-0"
bool           ::= "true" | "false"
symbol         ::= [a-z] [a-z0-9-]*
string         ::= '"' schar* '"'
digit          ::= [0-9]
hexdigit       ::= [0-9a-f]
schar          ::= unescaped | escape
unescaped      ::= <any Unicode scalar value other than
                  U+0022, U+005C, U+000A, U+0009>
escape         ::= U+005C U+0022    ; the two characters \"
               | U+005C U+005C    ; the two characters \\
               | U+005C "n"       ; the two characters \n
               | U+005C "t"       ; the two characters \t
\end{lstlisting}

Observe what does \emph{not} appear: no offset, no compressed length, no
compression algorithm, no chunk id, no content hash, no blob id. Every one is
either physical or derivable, and
Requirement~\ref{req:textproj:derive-or-carry} forbids both. The lone opaque
identity the grammar carries is the \texttt{SnapshotId}.

\textbf{Every production is expanded.} Version 0.1.0 left \texttt{kind},
\texttt{action}, \texttt{policy}, \texttt{constraints}, and \texttt{barrier}
derived-but-unwritten, and said so; that gap is closed. Barriers and affected
object kinds are byte strings, because the bundle holds them opaquely and the
projection interprets nothing the bundle does not.

At exactly the \texttt{value} positions delimited by
Requirement~\ref{req:textproj:operation-vocabulary}, the one thing still
\emph{read} rather than restated is a Chapter-5 value's field list. That is
deliberate: Requirement~\ref{req:textproj:value-projection} is a rule applied to
the core specification's ratified listings, not a copy of them.
A rule cannot drift from what it reads. An implementation that projects a value's
fields in an order other than the declaration order disagrees with the core
specification, not with this document.

Note the operation-kind names are the \emph{Operation Catalog's} section names
(\texttt{create-region}, \texttt{create-staff}), not the \texttt{OperationKindTag}
names (\texttt{InsertRegion}, \texttt{InsertStaff}). The tag space renamed three
pairs for reasons of its own (Binary Format,
\sectionsc{\texttt{OperationKindTag}}); the projection follows the semantics, not
the tag.

% ===========================================================================
\chapter{A Worked Example}
\label{ch:example}

\emph{Non-normative.} The byte strings below are illustrative but
\emph{well-formed}: every one is a grammar-valid \texttt{bytes} terminal, because
an example that cannot be parsed teaches the wrong lesson. The conformance vectors
that pin \emph{real} bytes are a deliverable of the implementation.

A document of one operation --- a transposition of two pitches up a perfect
fifth, over a compacted base --- projects to five lines:

\begin{lstlisting}
(text-projection (0 7 0))
(document #x05050505050505050505050505050505)
(profile full (0 1 0) (constraints 67108864 (retention 1 () true)))
(canonical-base #x1f8b0000000000000000000000000000 #x 1 full (schema 0 1) #x0000)
(envelope #x00000000000000070000000000000001 #x00000000000000000000000011223344 (stamp 42 7 #x00000000000000070000000000000001) (causal ((#x0000000000000001 3)) (#x00000000000000020000000000000009)) (some #x00000000000000070000000000000005) (primitive (transpose-interval (#x00000000000000070000000000000001 #x00000000000000070000000000000002) (transposition-interval 4 7))))
\end{lstlisting}

The envelope's targets are a \emph{set}: strictly increasing, no duplicates
(Operation Catalog, \texttt{req:opcat:transpose-interval-targets}; Binary Format
$\mathrm{seq}^{\Uparrow}$). A parser \MUST{} reject a duplicate rather than
absorb it, exactly as the binary decoder does.

% ===========================================================================
\chapter{Revision History}
\label{ch:history}

\begin{longtable}{p{2cm} p{2.5cm} p{9cm}}
  \toprule
  \textbf{Date} & \textbf{Section} & \textbf{Change} \\
  \midrule
  \endhead
  \today & All & 0.1.0 --- Initial companion. Supplies the canonical
  s-expression form the core specification's Chapter~8
  \sectionsc{Text Projection} declares normative and leaves unwritten, and which
  the Binary Format companion excludes as ``the \emph{Text Projection}
  companion's''. Ratified: reduced state is preserved by \emph{determining} it,
  never by a second literal copy (\texttt{req:textproj:reduced-state-derived});
  a canonical base snapshot is inlined as one opaque byte string, because a
  pruned document's base is derivable from nothing and a reference-only
  projection would be lossy (\texttt{req:textproj:base-snapshot-inline});
  lowercase hex is the single byte-string encoding
  (\texttt{req:textproj:hex}); one envelope per line, so a line-based merge
  conflict is exactly an envelope conflict
  (\texttt{req:textproj:envelope-per-line}). Parsing is strict-canonical
  (\texttt{req:textproj:strict-parse}) --- normalizing non-canonical text is
  accepting it --- and conformance requires \emph{both} directions
  (\texttt{req:textproj:conformance}). No implementation yet; this document is
  the design gate. \\
    \today & Chapters 3, 5 & 0.2.0 --- Canonical-manifest coverage. 0.1.0 was
  \emph{lossy for documents that are valid today}, and its claim to preserve the
  manifest's canonical roots was false in three ways: a canonical blob (an
  embedded image, font, or recording referenced by a canonical operation) had no
  representation at all; an \texttt{ExtensionDeclaration} lost its semantic
  version and its affected object kinds, and left its preserved chunk roots
  undefined; and \texttt{ProfileId::Custom} was unrepresentable, a symbol being
  required where a sixteen-byte registry id is carried.

  The three share one cause, now stated as
  \texttt{req:textproj:derive-or-carry}: a \texttt{ChunkRef} and a
  \texttt{BlobRef} are \emph{physical} references --- offset, compressed length,
  compression --- which the projection may not preserve, and they also carry
  \emph{derivable} identities, which it may not duplicate. Carry the content and
  the semantic attributes; re-derive the rest. The sole non-derivable identity in
  schema major~0 is \texttt{SnapshotId}, which the Binary Format companion pins
  as opaque and forbids readers to derive.

  Consequently: \texttt{req:textproj:canonical-blobs} (a canonical blob is
  projected; a non-canonical one is not), \texttt{req:textproj:profile-id}
  (\texttt{(custom \#x...)}), \texttt{req:textproj:extension-declaration}
  (every field, chunks as kind + schema + payload), and
  \texttt{req:textproj:base-snapshot-inline} extended to say what the inlined
  payload \emph{is} and how the root chunk id and the snapshot hash are
  re-derived rather than read. The core specification's own list of what the
  projection preserves omitted canonical blobs; it is corrected there too. The
  0.1.0 ratifications --- reduced state derived, base inlined, hex, one envelope
  per line, strict parsing --- stand unchanged. \\
    \today & Chapters 3, 5 & 0.3.0 --- Every production expanded. 0.1.0 left
  \texttt{kind}, \texttt{action}, \texttt{policy}, \texttt{constraints} and
  \texttt{barrier} derived-but-unwritten and admitted it; all are now written.
  Barriers and affected object kinds are byte strings, because the bundle holds
  them opaquely and the projection interprets nothing the bundle does not.

  Operation payloads embed Chapter-5 canonical values, and those are projected by
  \emph{one rule} rather than by forty productions
  (\texttt{req:textproj:value-projection}): a struct is
  \texttt{(<type-name> <field>\ldots)} with fields positional in the ratified
  declaration order; a newtype is transparent, as it is in the binary form; a
  tagged union is \texttt{(<variant> <field>\ldots)}; an option is \texttt{()}
  or \texttt{(some v)}; a sequence keeps the binary form's order. Restating the
  Chapter-5 model here would have put two normative listings on one struct, which
  is the drift P13-I1 was opened to close --- and a rule cannot drift from what it
  reads.

  Two leaf decisions follow from canonicality rather than taste. A rational is
  \texttt{(ratio n d)} in lowest terms with the sign on the numerator. A
  \texttt{CanonicalF64} is the byte string of its eight canonical IEEE~754 bytes,
  never a decimal: decimal float text is not canonically unique, so a decimal
  tempo would break \texttt{req:textproj:canonical-text} at the first tempo mark.

  Operation-kind names are the Operation Catalog's section names, not the
  \texttt{OperationKindTag} names, which renamed three pairs for reasons of the
  tag space. Still no implementation. \\
    \today & Chapters 3, 5 & 0.4.0 --- Two normative corrections found in review.

  \emph{The grammar contradicted its own escape requirement.}
  \texttt{req:textproj:string-escapes} obliges a writer to escape the backslash
  and a parser to reject a bare one, while \texttt{unescaped} admitted it. The
  escape productions are now spelled out as two-character sequences and
  \texttt{unescaped} excludes U+0022, U+005C, U+000A, and U+0009 by codepoint.
  Both characters of an escape are written as codepoints where they are the
  delimiter or the introducer: a quoted terminal for the backslash reads as
  \emph{two} backslashes and would make every escape three characters long. For
  the same reason the mono font no longer applies TeX ligatures, which rendered
  U+0022 as a right curly quote and \texttt{-{}-} as an en dash --- a document
  that specifies a text syntax must not misprint it.

  \emph{``Keep the binary order'' does not work for every sequence.} It is
  available only where the binary order reads preserved data, and two sequences
  fail that test: \texttt{blob\_roots} is sorted by the full \texttt{BlobRef}
  encoding, which contains the offset, the compressed length, and the compression;
  and an extension's preserved chunk roots are sorted by \texttt{ChunkRef}'s
  order, whose key is kind, then content hash, then \emph{offset}. Under the
  blanket rule, relocating a chunk --- which changes no semantics --- would have
  changed the text, and two entries indistinguishable after erasure would have
  produced duplicate lines. \texttt{req:textproj:derived-ordering} orders and
  de-duplicates those two sequences by their \emph{projected form}. Every other
  sequence keeps the binary order: the profile and extension declarations sort on
  semantic \texttt{(id, version)} keys, and the envelopes on canonical operation
  order.

  Also: a committed grammar-completeness test now checks that no nonterminal is
  undefined or unreachable, that the escape rule excludes the four codepoints and
  admits exactly the four two-character sequences, that the mono font substitutes
  no glyphs, and that the operation-kind and chunk-kind productions are exactly
  the tag vocabularies --- derived from \texttt{OperationKindTag} and
  \texttt{ChunkKind}, not transcribed. Every locator in it finds its production by
  name; the checks it replaced were anchored to a column, and a reflow would have
  silently switched them off. The 0.3.0 claim of a ``machine-checked'' grammar was
  true of one run and of nothing durable. Still no implementation. \\
    \today & Chapters 3, 5 & 0.5.0 --- Found by starting the implementation: the
  grammar could not derive an ordinary pitched note.

  \emph{The projection is schema-directed}
  (\texttt{req:textproj:schema-directed}), and now says so. \texttt{value} had no
  alternative for a sequence at all, though clause~5 required one; and once
  added, a sequence whose first element is a fieldless variant is
  shape-identical to a struct, while \texttt{()} is both the empty sequence and
  the absent option. Both collisions are reachable from the first pitched note in
  any score --- \texttt{PitchedEvent} carries \texttt{articulations} and
  \texttt{ornaments} over zero-field marks, beside an optional
  \texttt{DynamicMark}. The collision is irreducible without new syntax, and new
  syntax would buy nothing:
  \texttt{req:textproj:strict-parse} already obliges a parser to reject a
  duplicate in a set-typed field, which it cannot do without the schema. So
  \texttt{value} now collapses to \texttt{"(" value* ")"} or a leaf --- all shape
  can honestly say --- and the requirement assigns meaning by expected type.

  \emph{Three consequences, stated.} A struct with no fields is the bare symbol,
  as a fieldless variant is. A byte string is not a sequence: where the binary
  form writes a length-prefixed run of bytes, so does the projection, never a
  list of integers. And the grammar's repetitions now carry a notation rule ---
  adjacent elements are separated by exactly one space --- without which
  \texttt{"(transpose (" bytes* ") "} spelled two targets as one run of hex.

  Still no implementation; this is what the first day of writing it found. \\
    \today & Chapters 3, 5, 6 & 0.6.0 --- The operation vocabulary is
  grammar-directed, and the boundary is now normative
  (\texttt{req:textproj:operation-vocabulary}). The envelope, stamp, causal
  context, payload, operation kinds, and their sub-vocabularies follow the
  grammar's productions; \texttt{req:textproj:value-projection} applies exactly
  at the \texttt{value} positions those productions name. Operation-kind payload
  records are inlined because they name variant types but add neither a binary
  wrapper nor a modelling distinction.

  \texttt{transpose-interval} now takes its target byte strings followed by one
  \texttt{value}. Its \texttt{TranspositionInterval} therefore has the same
  \texttt{(transposition-interval ...)} spelling it has at every other value
  position, and the grammar has no special \texttt{interval} production. The
  committed grammar gate locks the requirement, its reliance-point citations,
  and that absence. \\
    \today & Chapters 3, 5 & 0.7.0 --- Canonical blob rejection and
  single-version header gating. At this version no canonical operation or
  canonical reduced state can reference a \texttt{BlobId}; every
  \texttt{(blob ...)} line is therefore unreferenced and non-canonical, and a
  parser must reject it
  (\texttt{req:textproj:reject-unreferenced-blobs}). Accepting it would stage
  content the next projection silently drops, lose data, and falsify the
  text-to-binary-to-text round trip. Forward compatibility belongs to
  header-version gating, not lenient blob acceptance.

  The parser accepts exactly the implemented companion header,
  \texttt{(0 7 0)}, and rejects every other version
  (\texttt{req:textproj:header-version}). Multi-version acceptance and text
  migrate-on-read remain deferred, in the same posture as op-payload
  migrate-on-read: a future real consumer must bring an explicit,
  version-keyed migration path rather than teaching the current parser to
  speculate. The worked example now uses the implemented header and contains
  only grammar-valid, canonical lines. \\
\bottomrule
\end{longtable}

\end{document}
